---
title: "Supplementary Information 2 - Simulation studies"
subtitle: "Community performance variability as a predictor of community stability"
authors: "Francesco Polazzo, Til Hämmig, Paula Eileen Stöckli, Shyamolina Ghosh, Owen L. Petchey"
editor: source
execute:
  echo: false
  warning: false
  message: false
  cache: false
format:
  html:
    page-layout: full
    html-math-method: mathml
    toc: true
    toc-location: left
    toc-depth: 3
    toc-expand: true
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
    number-sections: true
crossref:
  fig-title: Figure S
  tbl-title: Table S
  fig-prefix: Figure S
  tbl-prefix: Table S
  custom:
    - kind: float
      key: suppfig
      latex-env: suppfig
      reference-prefix: Figure S
      caption-prefix: Figure S
      space-before-numbering: false

    - kind: float
      key: supptbl
      latex-env: supptbl
      reference-prefix: Table S
      caption-prefix: Table S
      space-before-numbering: false
---



```{r setup}
library(dplyr)
library(forcats)
library(ggdist)
library(ggplot2)
library(knitr)
library(patchwork)
library(purrr)
library(ranger)
library(tibble)
library(tidyr)
library(viridisLite)

experiment_root <- "/Users/francesco/Documents/GitHub/CPCs_predict_CommunityStability/publication_experiments/data"
selected_experiments <- c("LV1", "LV2", "CR1", "CR2")

`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0) y else x
}

theme_figure <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      strip.text = element_text(face = "bold"),
      legend.position = "bottom"
    )
}

theme_concept <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      legend.position = "none",
      plot.title = element_text(hjust = 0.5, size = base_size),
      plot.subtitle = element_text(hjust = 0.5, size = base_size - 1)
    )
}

theme_clean <- function(base_size = 12) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", colour = "black"),
      legend.position = "bottom",
      plot.tag = element_text(face = "bold")
    )
}

format_experiment_label <- function(x) {
  recode(
    x,
    LV1 = "LV 1",
    LV2 = "LV 2",
    CR1 = "CR 1",
    CR2 = "CR 2"
  )
}

case_spec_value <- function(case_spec, path, default = NA) {
  parts <- strsplit(path, ".", fixed = TRUE)[[1]]
  value <- case_spec
  for (part in parts) {
    if (!is.list(value) || is.null(value[[part]])) {
      return(default)
    }
    value <- value[[part]]
  }
  if (length(value) != 1) {
    return(default)
  }
  value
}

case_spec_numeric <- function(data, path, default = NA_real_) {
  vapply(
    data$case_spec,
    function(spec) as.numeric(case_spec_value(spec, path, default)),
    numeric(1)
  )
}

case_spec_character <- function(data, path, default = NA_character_) {
  vapply(
    data$case_spec,
    function(spec) as.character(case_spec_value(spec, path, default)),
    character(1)
  )
}

nominal_lv_strength <- function(labels) {
  parsed <- sub(".*(?:mean|median)_([0-9.]+).*", "\\1", labels, perl = TRUE)
  parsed[parsed == labels] <- NA_character_
  as.numeric(parsed)
}

read_experiment <- function(experiment) {
  path <- file.path(experiment_root, experiment, "community_measures.RDS")
  readRDS(path) |>
    as_tibble() |>
    mutate(
      experiment = experiment,
      framework = if_else(grepl("^LV", experiment), "Lotka-Volterra", "Consumer-resource")
    )
}

standardise_experiment <- function(experiment) {
  dat <- read_experiment(experiment)

  if (grepl("^LV", experiment)) {
    labels <- case_spec_character(dat, "interactions.selected.label")
    dat |>
      mutate(
        treatment_label = labels,
        treatment_value = nominal_lv_strength(labels),
        treatment_plot = as.character(round(treatment_value, 2)),
        treatment_axis = "Nominal interaction strength",
        design_mean_optimum = case_spec_numeric(dat, "traits.birth_optimum.mean"),
        design_optimum_range = case_spec_numeric(dat, "traits.birth_optimum.range"),
        design_width_mean = case_spec_numeric(dat, "traits.birth_width.mean"),
        design_width_range = case_spec_numeric(dat, "traits.birth_width.range"),
        raw_cpc = as.numeric(cv_community_performance_info),
        raw_cpc_null = as.numeric(cv_community_performance_naive),
        viability_binary_cpc = as.numeric(cv_community_viability_binary_info),
        viability_binary_cpc_null = as.numeric(cv_community_viability_binary_naive),
        actual_cpc = as.numeric(cv_community_performance_info),
        null_cpc = as.numeric(cv_community_performance_naive)
      )
  } else {
    private_use <- case_spec_numeric(dat, "resources.private_use.mean")
    dat |>
      mutate(
        private_use_mean = private_use,
        treatment_value = 1 - private_use_mean,
        treatment_value = case_when(
          treatment_value <= 0.025 ~ 0,
          treatment_value >= 0.975 ~ 1,
          TRUE ~ treatment_value
        ),
        treatment_plot = as.character(round(treatment_value, 2)),
        treatment_label = paste0("shared_resource_", treatment_plot),
        treatment_axis = "Shared-resource use",
        design_mean_optimum = case_spec_numeric(dat, "traits.uptake_optimum.mean"),
        design_optimum_range = case_spec_numeric(dat, "traits.uptake_optimum.range"),
        design_width_mean = case_spec_numeric(dat, "traits.uptake_width.mean"),
        design_width_range = case_spec_numeric(dat, "traits.uptake_width.range"),
        raw_cpc = as.numeric(cv_community_performance_info),
        raw_cpc_null = as.numeric(cv_community_performance_naive),
        viability_binary_cpc = as.numeric(cv_community_viability_binary_info),
        viability_binary_cpc_null = as.numeric(cv_community_viability_binary_naive),
        actual_cpc = as.numeric(cv_community_viability_binary_info),
        null_cpc = as.numeric(cv_community_viability_binary_naive)
      )
  }
}

raw_data <- map_dfr(selected_experiments, standardise_experiment)

analysis_data <- raw_data |>
  mutate(
    stability = as.numeric(cv_total_abundance),
    distance_mean_optimum = abs(20 - as.numeric(realized_mean_performance_optimum)),
    sd_optimum = as.numeric(realized_sd_performance_optimum),
    framework_label = recode(
      framework,
      "Lotka-Volterra" = "Lotka-Volterra\nLV 1 + LV 2",
      "Consumer-resource" = "Consumer-resource\nCR 1 + CR 2"
    ),
    framework_label = factor(
      framework_label,
      levels = c("Lotka-Volterra\nLV 1 + LV 2", "Consumer-resource\nCR 1 + CR 2")
    ),
    experiment_label = factor(
      format_experiment_label(experiment),
      levels = format_experiment_label(selected_experiments)
    ),
    treatment_plot = factor(
      treatment_plot,
      levels = c("0", "0.25", "0.5", "0.75", "1")
    )
  ) |>
  filter(
    is.finite(stability),
    is.finite(actual_cpc),
    is.finite(null_cpc),
    is.finite(distance_mean_optimum),
    is.finite(sd_optimum),
    is.finite(richness),
    is.finite(treatment_value),
    !is.na(env_series_id)
  )

experiment_audit <- analysis_data |>
  group_by(experiment, framework) |>
  summarise(
    cases = n(),
    treatment_levels = n_distinct(treatment_plot),
    temperature_series = n_distinct(env_series_id),
    richness = paste(sort(unique(richness)), collapse = ", "),
    optimum_means = paste(sort(unique(design_mean_optimum)), collapse = ", "),
    optimum_ranges = paste(sort(unique(design_optimum_range)), collapse = ", "),
    width_means = paste(sort(unique(design_width_mean)), collapse = ", "),
    width_ranges = paste(sort(unique(design_width_range)), collapse = ", "),
    gamma = paste(sort(unique(one_over_f_gamma)), collapse = ", "),
    .groups = "drop"
  )
```

# Overview

This supplementary information presents a reproducible record of
the simulations used in the manuscript entitled: Community performance curves 
predict temporal stability despite interaction effects.
It contains the conceptual construction of community performance ($CP$),
the simulation experiments, the
generalisation of $CP$ through model-specific link functions, and additional
simulation analyses.

The simulation section uses two Lotka--Volterra runs and two consumer-resource
runs, reported as LV 1, LV 2, CR 1, and CR 2. These runs were designed to test
whether temporal variability in community performance predicts temporal
variability in total abundance across interaction structures and modelling
frameworks.

@fig-conceptual-cp shows the raw $CP$ concept. @fig-viability-cp gives the
conceptual construction of viability $CP$. @fig-main-simulation-results gives the
main simulation result and associated random-forest analysis. Later sections
define the model-specific $CP$ construction and compare synchrony predicted from
species thermal performance curves with synchrony realised in the simulations.

# Simulation Studies

The simulation analyses use four factorial experiments. LV 1
and LV 2 are Lotka--Volterra simulations in which temperature-dependent
performance affects intrinsic growth. CR 1 and CR 2 are consumer-resource
simulations in which temperature-dependent uptake and resource sharing generate
abundance dynamics through resource-mediated competition. In all simulations,
the response variable is temporal variability of total abundance, measured as
the coefficient of variation of total abundance.

The main linear models were fitted within each simulation run,
interaction-strength or resource-sharing treatment, and temperature series. 

## Implementation

Simulation experiments were implemented with the R package
[`community.simulator`](https://opetchey.github.io/community.simulator/index.html),
which provides tools for specifying factorial simulation designs, generating
temperature time series, simulating temperature-dependent community dynamics,
and calculating community-level summary measures. The workflow used the package
functions for reading experiment specifications, creating experiment tables,
generating environmental time series, running dynamical simulations, and
extracting community measures. The package documentation provides the relevant
implementation details in the articles on
[YAML experiment templates](https://opetchey.github.io/community.simulator/articles/yaml_templates.html),
[getting started with a small experiment](https://opetchey.github.io/community.simulator/articles/experiment_getting_started.html),
and the model-specific walkthroughs for
[continuous-time Lotka--Volterra](https://opetchey.github.io/community.simulator/articles/lv_continuous_single_community.html)
and
[consumer-resource](https://opetchey.github.io/community.simulator/articles/consumer_resource_single_community.html)
simulations. Function-level documentation is provided for
[`run_experiment()`](https://opetchey.github.io/community.simulator/reference/run_experiment.html),
[`simulate_dynamics_from_spec()`](https://opetchey.github.io/community.simulator/reference/simulate_dynamics_from_spec.html),
[`get_community_measures_from_spec()`](https://opetchey.github.io/community.simulator/reference/get_community_measures_from_spec.html),
[`simulator_lv_continuous()`](https://opetchey.github.io/community.simulator/reference/simulator_lv_continuous.html),
and
[`simulator_consumer_resource_continuous()`](https://opetchey.github.io/community.simulator/reference/simulator_consumer_resource_continuous.html).

```{r}
#| label: tbl-simulation-audit
#| tbl-cap: "Publication simulation data included in the supplementary report."
#| tbl-cap-location: bottom

kable(
  experiment_audit
)
```

## Conceptual Construction of $CP$

```{r conceptual-figure-preparation, include=FALSE}
make_reddened_temperature <- function(seed) {
  n <- 35
  gamma <- 0.8
  target_mean <- 22.5
  target_sd <- 3.5
  temp_step <- 0.1

  set.seed(seed)
  f <- seq(1, n / 2)
  scales <- 1 / (f^(gamma / 2))
  phases <- runif(length(f), 0, 2 * pi)
  coeffs <- scales * exp(1i * phases)
  full_coeffs <- c(0, coeffs, rev(Conj(coeffs[-length(coeffs)])))
  raw_noise <- Re(fft(full_coeffs, inverse = TRUE))[1:n]
  temperature_reddened <- as.vector(scale(raw_noise)) * target_sd + target_mean
  temperature_reddened <- round(temperature_reddened / temp_step) * temp_step

  tibble(
    time = seq_len(n),
    temp = temperature_reddened
  ) |>
    drop_na()
}

df_red2 <- make_reddened_temperature(113)
experimental_day_index <- c(1:5, 8:12, 15:19, 22:26, 29:33)
T_norm2 <- df_red2$temp[experimental_day_index]

scale_temperature_trace <- function(temperature_df, y_values) {
  y_range <- range(y_values, na.rm = TRUE, finite = TRUE)
  temp_range <- range(temperature_df$temperature, na.rm = TRUE, finite = TRUE)

  if (!all(is.finite(y_range)) || diff(y_range) == 0 ||
      !all(is.finite(temp_range)) || diff(temp_range) == 0) {
    return(mutate(temperature_df, temperature_scaled = mean(y_range)))
  }

  temperature_df |>
    mutate(
      temperature_scaled = y_range[1] +
        (temperature - temp_range[1]) / diff(temp_range) * diff(y_range)
    )
}
```

```{r}
#| label: fig-conceptual-cp
#| fig-cap: "Conceptual construction of raw community performance ($CP$) using four idealised communities with widely spaced, tightly clustered, cool-skewed, or bimodal species optima. The first column shows species-level performance curves and the summed $CP$ curve. The second column shows null $CP$ time series under uniformly sampled temperatures. The third column shows actual $CP$ time series under an experienced temperature sequence. Dashed grey lines in the time-series panels show the temperature sequence rescaled to the plotting height."
#| fig-width: 15
#| fig-height: 12
#| fig-cap-location: bottom

concept_species_colours <- setNames(
  viridisLite::magma(5, begin = 0.20, end = 0.85),
  paste0("sp", 1:5)
)

concept_temperature_grid <- seq(10, 30, length.out = 250)

concept_null_temperatures <- tibble(
  time = seq_len(25),
  temperature = seq(10, 30, length.out = 25)
)

concept_actual_temperatures <- tibble(
  time = seq_along(T_norm2),
  temperature = T_norm2
)

concept_species_sets <- list(
  "Widely spaced optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(12, 15, 20, 25, 28),
    width = c(3, 5, 4, 6, 3)
  ),
  "Tightly clustered optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(19, 19.5, 20, 20.5, 21),
    width = c(3.5, 3.8, 4, 3.6, 3.9)
  ),
  "Cool-skewed optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(11, 12, 13, 14, 15),
    width = c(3, 4, 5, 3, 4)
  ),
  "Bimodal optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(11, 12, 13, 27, 29),
    width = c(3, 4, 5, 3, 4)
  )
)

concept_performance <- function(temperature, optimum, width) {
  exp(-0.5 * ((temperature - optimum) / width)^2)
}

concept_species_performance <- function(species_df, temperature_df) {
  crossing(
    temperature_df,
    species_df
  ) |>
    mutate(
      performance = concept_performance(temperature, Topt, width)
    )
}

concept_cp_timeseries <- function(species_df, temperature_df) {
  concept_species_performance(species_df, temperature_df) |>
    group_by(time) |>
    mutate(community_performance = sum(performance)) |>
    ungroup()
}

concept_viability_timeseries <- function(species_df, temperature_df, threshold = 0.35) {
  concept_species_performance(species_df, temperature_df) |>
    mutate(viable = as.numeric(performance >= threshold)) |>
    group_by(time) |>
    mutate(community_viability = mean(viable)) |>
    ungroup()
}

concept_curve_column <- function(species_df, community_label, show_title = FALSE, show_x = FALSE) {
  curve_temperatures <- tibble(
    time = seq_along(concept_temperature_grid),
    temperature = concept_temperature_grid
  )

  species_curves <- concept_species_performance(species_df, curve_temperatures)

  cp_curve <- species_curves |>
    group_by(temperature) |>
    summarise(community_performance = sum(performance), .groups = "drop")

  p <- ggplot() +
    geom_line(
      data = species_curves,
      aes(x = temperature, y = performance, colour = species),
      linewidth = 1.05,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_curve,
      aes(x = temperature, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    theme_clean(12) +
    labs(
      title = if (show_title) expression(paste("Species curves and ", CP)) else NULL,
      subtitle = community_label,
      x = "Temperature (°C)",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_timeseries_column <- function(species_df, temperature_df, title, show_x = FALSE) {
  plot_data <- concept_cp_timeseries(species_df, temperature_df)
  cp_data <- plot_data |>
    distinct(time, community_performance)

  cp_cv <- sd(cp_data$community_performance) / mean(cp_data$community_performance)
  temperature_trace <- scale_temperature_trace(
    temperature_df,
    c(plot_data$performance, cp_data$community_performance)
  )

  p <- ggplot() +
    geom_line(
      data = temperature_trace,
      aes(x = time, y = temperature_scaled),
      inherit.aes = FALSE,
      colour = "grey35",
      linewidth = 0.7,
      linetype = "22",
      alpha = 0.45
    ) +
    geom_line(
      data = plot_data,
      aes(x = time, y = performance, colour = species, group = species),
      linewidth = 0.95,
      alpha = 0.78
    ) +
    geom_point(
      data = plot_data,
      aes(x = time, y = performance, colour = species),
      size = 1.4,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_data,
      aes(x = time, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    annotate(
      "label",
      x = Inf,
      y = Inf,
      hjust = 1.03,
      vjust = 1.12,
      label = paste0("CV(CP) == ", sprintf("%.2f", cp_cv)),
      linewidth = 0.18,
      size = 3.4,
      parse = TRUE
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    theme_clean(12) +
    labs(
      title = title,
      x = "Time step",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_figure_rows <- imap(concept_species_sets, function(species_df, community_label) {
  row_index <- match(community_label, names(concept_species_sets))
  show_title <- row_index == 1
  show_x <- row_index == length(concept_species_sets)

  concept_curve_column(
    species_df,
    community_label,
    show_title = show_title,
    show_x = show_x
  ) |
    concept_timeseries_column(
      species_df,
      concept_null_temperatures,
      title = if (show_title) expression(atop(paste("Null ", CP), "over uniform temperatures")) else NULL,
      show_x = show_x
    ) |
    concept_timeseries_column(
      species_df,
      concept_actual_temperatures,
      title = if (show_title) expression(atop(paste("Actual ", CP), "over experienced temperatures")) else NULL,
      show_x = show_x
    )
})

wrap_plots(concept_figure_rows, ncol = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "none")
```

### Viability $CP$

```{r}
#| label: fig-viability-cp
#| fig-cap: "Conceptual construction of viability $CP$. Species performance curves are first transformed into binary viability using the dashed threshold. Community viability $CP$ is the fraction of species above that threshold at each temperature or time step. Dashed grey traces in the time-series panels show the temperature sequence rescaled to the panel height, and black curves show community-level viability $CP$."
#| fig-width: 15
#| fig-height: 12
#| fig-cap-location: bottom

viability_threshold <- 0.35

concept_viability_curve_column <- function(species_df, community_label, show_title = FALSE, show_x = FALSE) {
  curve_temperatures <- tibble(
    time = seq_along(concept_temperature_grid),
    temperature = concept_temperature_grid
  )

  species_curves <- concept_species_performance(species_df, curve_temperatures)

  viability_curve <- species_curves |>
    mutate(viable = as.numeric(performance >= viability_threshold)) |>
    group_by(temperature) |>
    summarise(community_viability = mean(viable), .groups = "drop")

  p <- ggplot() +
    geom_line(
      data = species_curves,
      aes(x = temperature, y = performance, colour = species),
      linewidth = 1.05,
      alpha = 0.70
    ) +
    geom_hline(
      yintercept = viability_threshold,
      colour = "grey35",
      linewidth = 0.55,
      linetype = "22"
    ) +
    geom_step(
      data = viability_curve,
      aes(x = temperature, y = community_viability),
      linewidth = 1.55,
      colour = "black"
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    coord_cartesian(ylim = c(0, 1.05)) +
    theme_clean(12) +
    labs(
      title = if (show_title) expression(paste("Species curves and viability ", CP)) else NULL,
      subtitle = community_label,
      x = "Temperature (°C)",
      y = "Performance / fraction viable",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_viability_timeseries_column <- function(species_df, temperature_df, title, show_x = FALSE) {
  plot_data <- concept_viability_timeseries(
    species_df,
    temperature_df,
    threshold = viability_threshold
  )

  viability_data <- plot_data |>
    distinct(time, community_viability)

  viability_cv <- sd(viability_data$community_viability) /
    mean(viability_data$community_viability)

  temperature_trace <- scale_temperature_trace(
    temperature_df,
    c(plot_data$viable, viability_data$community_viability)
  )

  p <- ggplot() +
    geom_line(
      data = temperature_trace,
      aes(x = time, y = temperature_scaled),
      inherit.aes = FALSE,
      colour = "grey35",
      linewidth = 0.7,
      linetype = "22",
      alpha = 0.45
    ) +
    geom_line(
      data = plot_data,
      aes(x = time, y = viable, colour = species, group = species),
      linewidth = 0.85,
      alpha = 0.65
    ) +
    geom_point(
      data = plot_data,
      aes(x = time, y = viable, colour = species),
      size = 1.4,
      alpha = 0.70
    ) +
    geom_step(
      data = viability_data,
      aes(x = time, y = community_viability),
      linewidth = 1.55,
      colour = "black"
    ) +
    annotate(
      "label",
      x = Inf,
      y = Inf,
      hjust = 1.03,
      vjust = 1.12,
      label = paste0("CV(CP) == ", sprintf("%.2f", viability_cv)),
      linewidth = 0.18,
      size = 3.4,
      parse = TRUE
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    coord_cartesian(ylim = c(-0.04, 1.05)) +
    theme_clean(12) +
    labs(
      title = title,
      x = "Time step",
      y = "Viability / fraction viable",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

viability_figure_rows <- imap(concept_species_sets, function(species_df, community_label) {
  row_index <- match(community_label, names(concept_species_sets))
  show_title <- row_index == 1
  show_x <- row_index == length(concept_species_sets)

  concept_viability_curve_column(
    species_df,
    community_label,
    show_title = show_title,
    show_x = show_x
  ) |
    concept_viability_timeseries_column(
      species_df,
      concept_null_temperatures,
      title = if (show_title) expression(atop(paste("Null viability ", CP), "uniform temperatures")) else NULL,
      show_x = show_x
    ) |
    concept_viability_timeseries_column(
      species_df,
      concept_actual_temperatures,
      title = if (show_title) expression(atop(paste("Actual viability ", CP), "experienced temperatures")) else NULL,
      show_x = show_x
    )
})

wrap_plots(viability_figure_rows, ncol = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "none")
```


```{r model-functions}
safe_r2 <- function(formula, data) {
  fit <- tryCatch(lm(formula, data = data), error = function(e) NULL)
  if (is.null(fit)) {
    return(NA_real_)
  }
  r2 <- summary(fit)$r.squared
  ifelse(is.finite(r2), r2, NA_real_)
}

grouped_r2 <- analysis_data |>
  group_by(experiment, framework, framework_label, treatment_plot, treatment_value, env_series_id) |>
  group_modify(~ {
    dat <- .x
    if (nrow(dat) < 4) {
      return(tibble(
        predictor = c("Actual CP variability", "Null CP variability", "Trait summaries"),
        r2 = NA_real_,
        n_cases = nrow(dat),
        note = "fewer than four observations in group"
      ))
    }

    tibble(
      predictor = c("Actual CP variability", "Null CP variability", "Trait summaries"),
      r2 = c(
        safe_r2(stability ~ actual_cpc, dat),
        safe_r2(stability ~ null_cpc, dat),
        safe_r2(stability ~ distance_mean_optimum * sd_optimum, dat)
      ),
      n_cases = nrow(dat),
      note = NA_character_
    )
  }) |>
  ungroup() |>
  mutate(
    predictor = factor(predictor, levels = c("Actual CP variability", "Null CP variability", "Trait summaries")),
    treatment_plot = factor(treatment_plot, levels = c("0", "0.25", "0.5", "0.75", "1"))
  )

summary_by_framework <- grouped_r2 |>
  group_by(framework, predictor) |>
  summarise(
    groups = sum(is.finite(r2)),
    median_r2 = median(r2, na.rm = TRUE),
    q25_r2 = quantile(r2, 0.25, na.rm = TRUE),
    q75_r2 = quantile(r2, 0.75, na.rm = TRUE),
    min_r2 = min(r2, na.rm = TRUE),
    max_r2 = max(r2, na.rm = TRUE),
    .groups = "drop"
  )

summary_by_treatment <- grouped_r2 |>
  group_by(framework, treatment_plot, predictor) |>
  summarise(
    median_r2 = median(r2, na.rm = TRUE),
    .groups = "drop"
  )

summary_by_experiment <- grouped_r2 |>
  group_by(experiment, framework, predictor) |>
  summarise(
    median_r2 = median(r2, na.rm = TRUE),
    .groups = "drop"
  )

get_summary_value <- function(framework_name, predictor_name, column = "median_r2") {
  summary_by_framework |>
    filter(framework == framework_name, predictor == predictor_name) |>
    pull({{ column }}) |>
    first()
}
```

## Simulation Results

Variability of $CP$ predicted community variability across both modelling
frameworks, but the relevant $CP$ scale differed between models. In the
Lotka--Volterra simulations, where temperature-dependent performance enters
directly through intrinsic growth, raw $CV(CP)$ explained a large fraction of
variation in total-abundance instability.
Actual $CP$ variability, $CV(CP_{\text{actual}})$, had a median grouped \(R^2\) of
`r round(get_summary_value("Lotka-Volterra", "Actual CP variability"), 2)`, and null
$CP$ variability, $CV(CP_{\text{null}})$, had a very similar median grouped \(R^2\) of
`r round(get_summary_value("Lotka-Volterra", "Null CP variability"), 2)`. By contrast, the
trait-summary model explained less variation, with a median grouped
\(R^2\) of `r round(get_summary_value("Lotka-Volterra", "Trait summaries"), 2)`.

In the consumer-resource simulations, raw uptake curves are filtered through
resource depletion, mortality, and competition for shared resources. Binary
viability therefore provides the $CP$ scale most directly related to total
abundance. Actual viability $CP$ variability, $CV(CP_{\text{actual}})$, had a median
grouped \(R^2\) of
`r round(get_summary_value("Consumer-resource", "Actual CP variability"), 2)`. Null
viability $CP$ variability, $CV(CP_{\text{null}})$, explained less variation
(\(R^2 = `r round(get_summary_value("Consumer-resource", "Null CP variability"), 2)`\)),
as did the trait-summary model
(\(R^2 = `r round(get_summary_value("Consumer-resource", "Trait summaries"), 2)`\)).
These results indicate that $CP$ variability is most informative when $CP$ is
constructed on a scale that corresponds to the abundance process being analysed.

The random-forest analysis gave a consistent ranking of predictors. Actual
$CP$ variability, $CV(CP_{\text{actual}})$, had the highest permutation importance in all four simulation runs. Null
$CP$ variability, $CV(CP_{\text{null}})$, had secondary permutation importance in the Lotka--Volterra runs and
lower permutation importance in the consumer-resource runs. Trait summaries,
richness, and interaction or resource-sharing treatment had lower permutation
importance once $CV(CP_{\text{actual}})$ was included.

```{r r2-summary-tables, include=FALSE}
kable(
  summary_by_framework |>
    mutate(across(ends_with("_r2"), ~ round(.x, 3))),
  caption = "Grouped linear-model explanatory power by modelling framework."
)

kable(
  summary_by_experiment |>
    mutate(median_r2 = round(median_r2, 3)),
  caption = "Median grouped explanatory power by experiment."
)
```

```{r random-forest, include=FALSE}
rf_variable_labels <- c(
  actual_cpc = "Actual CP variability",
  null_cpc = "Null CP variability",
  distance_mean_optimum = "|mean optimum - mean T|",
  sd_optimum = "SD optimum",
  richness = "Richness",
  treatment_numeric = "Interaction/resource sharing"
)

make_rf_data <- function(experiment_name) {
  analysis_data |>
    filter(experiment == experiment_name) |>
    mutate(treatment_numeric = as.numeric(treatment_value)) |>
    select(
      experiment,
      experiment_label,
      framework,
      stability,
      actual_cpc,
      null_cpc,
      distance_mean_optimum,
      sd_optimum,
      richness,
      treatment_numeric
    ) |>
    filter(
      if_all(
        c(stability, actual_cpc, null_cpc, distance_mean_optimum,
          sd_optimum, richness, treatment_numeric),
        is.finite
      )
    )
}

fit_rf_outputs <- function(experiment_name) {
  dat <- make_rf_data(experiment_name)
  if (nrow(dat) < 20) {
    empty_importance <- tibble(
        experiment = experiment_name,
        experiment_label = format_experiment_label(experiment_name),
        variable = names(rf_variable_labels),
        importance = NA_real_,
        scaled_importance = NA_real_,
        note = "too few cases for a stable random forest"
      )
    empty_predictions <- tibble(
      experiment = experiment_name,
      experiment_label = format_experiment_label(experiment_name),
      framework = NA_character_,
      observed = NA_real_,
      predicted = NA_real_,
      oob_r2 = NA_real_
    )
    return(list(importance = empty_importance, predictions = empty_predictions))
  }

  set.seed(123)
  rf <- ranger(
    stability ~ actual_cpc + null_cpc + distance_mean_optimum +
      sd_optimum + richness + treatment_numeric,
    data = dat,
    importance = "permutation",
    num.trees = 1000,
    mtry = 3
  )

  importance <- enframe(rf$variable.importance, name = "variable", value = "importance") |>
    mutate(
      experiment = experiment_name,
      experiment_label = format_experiment_label(experiment_name),
      scaled_importance = importance / max(importance, na.rm = TRUE),
      note = NA_character_
    )

  predictions <- dat |>
    transmute(
      experiment = experiment_name,
      experiment_label = first(experiment_label),
      framework = first(framework),
      observed = stability,
      predicted = rf$predictions,
      oob_r2 = rf$r.squared
    )

  list(importance = importance, predictions = predictions)
}

rf_outputs <- map(selected_experiments, fit_rf_outputs)

rf_importance <- map_dfr(rf_outputs, "importance") |>
  mutate(
    variable_label = recode(variable, !!!rf_variable_labels),
    variable_label = factor(variable_label, levels = unname(rf_variable_labels)),
    experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments))
  )

rf_prediction_data <- map_dfr(rf_outputs, "predictions") |>
  filter(is.finite(observed), is.finite(predicted)) |>
  mutate(
    experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments)),
    oob_label = paste0("OOB R² = ", round(oob_r2, 3))
  )

rf_oob_summary <- rf_prediction_data |>
  distinct(experiment, experiment_label, framework, oob_r2)

random_subset_r2 <- function(experiment_name) {
  dat <- make_rf_data(experiment_name)
  n_groups <- min(500L, max(1L, floor(nrow(dat) / 25L)))

  set.seed(100 + match(experiment_name, selected_experiments))
  dat |>
    mutate(random_group = sample(rep(seq_len(n_groups), length.out = n()))) |>
    group_by(random_group) |>
    summarise(
      n_pairs = sum(is.finite(stability) & is.finite(actual_cpc)),
      r2 = if (n_pairs >= 4) {
        cor(stability, actual_cpc, use = "complete.obs")^2
      } else {
        NA_real_
      },
      .groups = "drop"
    ) |>
    filter(is.finite(r2)) |>
    mutate(
      experiment = experiment_name,
      experiment_label = format_experiment_label(experiment_name)
    )
}

random_subset_data <- map_dfr(selected_experiments, random_subset_r2) |>
  mutate(experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments)))

kable(
  if (knitr::pandoc_to("docx")) {
    rf_importance |>
      select(experiment, variable = variable_label, scaled_importance) |>
      mutate(scaled_importance = round(scaled_importance, 3))
  } else {
    rf_importance |>
      select(experiment, variable = variable_label, importance, scaled_importance, note) |>
      mutate(across(c(importance, scaled_importance), ~ round(.x, 3)))
  },
  caption = "Random-forest permutation importance. Values are scaled within each experiment for plotting."
)
```

```{r}
#| label: fig-main-simulation-results
#| fig-cap: 'Explanatory power of community performance metrics across the four simulation experiments. Panel (a) shows linear-model R² values fitted within each run, treatment level, and temperature series; models were not grouped by mean thermal optimum. The x-axis gives nominal interaction strength for LV 1 and LV 2 and resource sharing for CR 1 and CR 2, where 0 indicates no shared resource and 1 indicates fully shared resource use. Facets show actual $CP$ variability ($CV(CP_{\text{actual}})$), null $CP$ variability ($CV(CP_{\text{null}})$), and trait summaries. For LV 1 and LV 2, actual and null $CP$ variability are based on raw community performance. For CR 1 and CR 2, actual and null $CP$ variability are based on binary viability $CP$. Trait summaries are the distance between realised mean optimum and mean environmental temperature, the realised standard deviation of optima, and their interaction. Panel (b) shows random-subset robustness for $CV(CP_{\text{actual}})$. Panel (c) shows random-forest permutation importance for the same predictor set, scaled within each run. Panel (d) compares observed instability with out-of-bag predicted instability.'
#| fig-width: 15
#| fig-height: 10.5
#| fig-cap-location: bottom

importance_offsets <- tibble(
  experiment_label = factor(
    format_experiment_label(selected_experiments),
    levels = format_experiment_label(selected_experiments)
  ),
  importance_offset = c(-0.24, -0.08, 0.08, 0.24)
)

importance_data <- rf_importance |>
  filter(is.finite(scaled_importance)) |>
  left_join(importance_offsets, by = "experiment_label") |>
  mutate(
    variable_label = factor(variable_label, levels = rev(unname(rf_variable_labels))),
    variable_y = as.numeric(variable_label) + importance_offset
  )

experiment_levels <- unique(c(
  as.character(random_subset_data$experiment_label),
  as.character(importance_data$experiment_label),
  as.character(rf_prediction_data$experiment_label)
))

experiment_colours <- setNames(
  magma(
    n = length(experiment_levels),
    begin = 0.20,
    end = 0.85
  ),
  experiment_levels
)

framework_levels <- c("LV", "CR")

framework_colours <- setNames(
  magma(
    n = length(framework_levels),
    begin = 0.20,
    end = 0.85
  ),
  framework_levels
)

random_subset_data <- random_subset_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

importance_data <- importance_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

rf_prediction_data <- rf_prediction_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

common_theme <- theme_figure(12) +
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13, colour = "black"),
    axis.text = element_text(size = 11, colour = "black"),
    strip.text = element_text(size = 12, face = "bold", colour = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.key.size = grid::unit(0.7, "cm"),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 7, r = 9, b = 7, l = 7)
  )

panel_a <- grouped_r2 |>
  mutate(
    framework_short = factor(
      framework,
      levels = c("Lotka-Volterra", "Consumer-resource"),
      labels = framework_levels
    )
  ) |>
  ggplot(aes(
    x = treatment_plot,
    y = r2,
    colour = framework_short,
    fill = framework_short
  )) +
  stat_halfeye(
    position = position_dodge(width = 0.65),
    adjust = 0.8,
    width = 0.85,
    slab_alpha = 0.40,
    point_interval = median_qi,
    .width = c(0.5, 0.8),
    point_size = 3
  ) +
  geom_point(
    position = position_jitterdodge(
      dodge.width = 0.65,
      jitter.width = 0.03
    ),
    alpha = 0.20,
    size = 2.2,
    show.legend = FALSE
  ) +
  facet_wrap(~ predictor, nrow = 1) +
  scale_colour_manual(values = framework_colours, name = "Framework", drop = FALSE) +
  scale_fill_manual(values = framework_colours, drop = FALSE, guide = "none") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Interaction strength / resource sharing",
    y = expression(paste("Grouped explanatory power (", R^2, ")"))
  ) +
  common_theme +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 11, angle = 0, hjust = 0.5),
    strip.text = element_text(size = 13, face = "bold")
  )

random_subset_panel <- random_subset_data |>
  ggplot(aes(x = r2, fill = experiment_label)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 32,
    colour = "white",
    linewidth = 0.25,
    alpha = 0.75
  ) +
  geom_density(linewidth = 0.9, colour = "grey20") +
  facet_wrap(~ experiment_label, ncol = 2) +
  scale_fill_manual(values = experiment_colours, drop = FALSE, guide = "none") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = expression(paste("Random-subset ", R^2, " for ", CV(CP[actual]))),
    y = "Density"
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 10.5),
    strip.text = element_text(size = 11.5, face = "bold")
  )

importance_plot <- importance_data |>
  ggplot(aes(
    y = variable_y,
    x = scaled_importance,
    colour = experiment_label,
    shape = experiment_label
  )) +
  geom_segment(
    aes(x = 0, xend = scaled_importance, yend = variable_y),
    linewidth = 0.65
  ) +
  geom_point(size = 2.8, stroke = 0.6) +
  scale_colour_manual(values = experiment_colours, drop = FALSE, name = "Run") +
  scale_shape_manual(values = c(16, 17, 15, 18), drop = FALSE, name = "Run") +
  scale_y_continuous(
    breaks = seq_along(levels(importance_data$variable_label)),
    labels = levels(importance_data$variable_label)
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Scaled permutation importance",
    y = "Predictor variables"
  ) +
  common_theme +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 11),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4)
  )

rf_accuracy_panel <- rf_prediction_data |>
  ggplot(aes(
    x = predicted,
    y = observed,
    colour = experiment_label
  )) +
  geom_point(alpha = 0.25, size = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey50",
    linewidth = 0.65
  ) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +
  geom_text(
    data = rf_prediction_data |> distinct(experiment_label, oob_label),
    aes(x = -Inf, y = Inf, label = oob_label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.3,
    size = 3.8,
    colour = "black"
  ) +
  facet_wrap(~ experiment_label, ncol = 2, scales = "free") +
  scale_colour_manual(values = experiment_colours, drop = FALSE, guide = "none") +
  labs(
    x = expression(paste("Out-of-bag predicted variability (", CV(CP[actual]), ")")),
    y = expression(paste("Observed variability (", plain(CV), "(", N[plain(tot)], "))"))
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 10.5),
    strip.text = element_text(size = 11.5, face = "bold")
  )

panel_a /
  (random_subset_panel | importance_plot | rf_accuracy_panel) +
  plot_layout(
    heights = c(1, 1.05),
    widths = c(1, 1, 1),
    guides = "keep"
  ) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )
```

```{r}
#| label: tbl-r2-summary-by-treatment
#| tbl-cap: "Mean grouped explanatory power by modelling framework, grouping scheme, and predictor. Treatment-level rows come from linear models fitted within each run, interaction-strength or resource-sharing treatment, and temperature series. Pooled-treatment rows show the corresponding grouped R² values when treatment levels are not used for grouping; these models are fitted within each run and temperature series only."
#| tbl-cap-location: bottom

r2_by_temperature_series <- analysis_data |>
  group_by(experiment, framework, framework_label, env_series_id) |>
  group_modify(~ {
    dat <- .x
    if (nrow(dat) < 4) {
      return(tibble(
        predictor = c("Actual CP variability", "Null CP variability", "Trait summaries"),
        r2 = NA_real_,
        n_cases = nrow(dat),
        note = "fewer than four observations in group"
      ))
    }

    tibble(
      predictor = c("Actual CP variability", "Null CP variability", "Trait summaries"),
      r2 = c(
        safe_r2(stability ~ actual_cpc, dat),
        safe_r2(stability ~ null_cpc, dat),
        safe_r2(stability ~ distance_mean_optimum * sd_optimum, dat)
      ),
      n_cases = nrow(dat),
      note = NA_character_
    )
  }) |>
  ungroup() |>
  mutate(
    predictor = factor(predictor, levels = c("Actual CP variability", "Null CP variability", "Trait summaries")),
    treatment_plot = "Pooled",
    grouping = "Run + temperature series"
  )

r2_summary_by_treatment_table <- bind_rows(
  grouped_r2 |>
    mutate(grouping = "Run + treatment + temperature series"),
  r2_by_temperature_series
) |>
  mutate(
    experiment_label = factor(
      format_experiment_label(experiment),
      levels = format_experiment_label(selected_experiments)
    )
  ) |>
  group_by(framework, experiment_label, grouping, treatment_plot, predictor) |>
  summarise(
    groups = sum(is.finite(r2)),
    mean_r2 = mean(r2, na.rm = TRUE),
    median_r2 = median(r2, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    mean_r2 = round(mean_r2, 3),
    median_r2 = round(median_r2, 3)
  ) |>
  arrange(framework, experiment_label, grouping, treatment_plot, predictor)

kable(
  r2_summary_by_treatment_table,
  align = c("l", "l", "l", "c", "l", "r", "r", "r")
)
```

## Simulation Interpretation

The simulations support two linked conclusions. First, variability of $CP$
predicts temporal community variability across both modelling frameworks, despite the fact
that the frameworks differ substantially in how species responses enter
abundance dynamics. Second, the appropriate $CP$ scale depends on the model. In
the Lotka--Volterra simulations, raw $CV(CP)$ is appropriate because
temperature-dependent performance enters directly into intrinsic growth. In the
consumer-resource simulations, abundance depends on resource supply, depletion,
mortality, and competition for shared resources. Binary viability $CP$ is
therefore the more relevant community response for predicting total abundance
variability.

The consumer-resource simulations address a limitation of the Lotka--Volterra
simulations. In the Lotka--Volterra model,
interaction coefficients are not temperature dependent. In the consumer-resource
model, uptake rates are temperature dependent and species compete through shared
resources, implying temperature-dependent effective interaction strengths. The
high predictive power of actual viability $CP$ variability, $CV(CP_{\text{actual}})$, in CR 1 and CR 2 therefore
indicates that the main $CP$-stability result is robust to a mechanistic setting
in which interaction strengths can vary with temperature.

## Constructing $CP$ on an Appropriate Scale

```{r cp-generalized-publication-data, include=FALSE}
publication_link_data <- bind_rows(
  analysis_data |>
    transmute(
      model_type = "Lotka--Volterra",
      framework,
      framework_label,
      experiment_label,
      link_function = "Identity link",
      CP_CV = raw_cpc,
      community_stability = 1 / stability,
      community_instability = stability
    ),
  analysis_data |>
    transmute(
      model_type = "Lotka--Volterra",
      framework,
      framework_label,
      experiment_label,
      link_function = "Binary link",
      CP_CV = viability_binary_cpc,
      community_stability = 1 / stability,
      community_instability = stability
    ),
  analysis_data |>
    transmute(
      model_type = "Consumer-resource",
      framework,
      framework_label,
      experiment_label,
      link_function = "Identity link",
      CP_CV = raw_cpc,
      community_stability = 1 / stability,
      community_instability = stability
    ),
  analysis_data |>
    transmute(
      model_type = "Consumer-resource",
      framework,
      framework_label,
      experiment_label,
      link_function = "Binary link",
      CP_CV = viability_binary_cpc,
      community_stability = 1 / stability,
      community_instability = stability
    )
) |>
  filter(is.finite(CP_CV), is.finite(community_stability), is.finite(community_instability)) |>
  mutate(
    model_type = factor(model_type, levels = c("Lotka--Volterra", "Consumer-resource")),
    link_function = factor(link_function, levels = c("Identity link", "Binary link"))
  )

publication_link_summaries <- publication_link_data |>
  group_by(model_type, link_function) |>
  summarise(
    n = n(),
    cp_cv_sd = sd(CP_CV),
    stability_sd = sd(community_stability),
    r_squared = ifelse(
      n > 2 && cp_cv_sd > 0 && stability_sd > 0,
      summary(lm(community_stability ~ CP_CV))$r.squared,
      NA_real_
    ),
    .groups = "drop"
  ) |>
  mutate(
    panel_label = ifelse(
      is.na(r_squared),
        paste0("n = ", n, "\nCV(CP) constant"),
      paste0("n = ", n, "\nR² = ", formatC(r_squared, format = "f", digits = 2))
    )
  )

plot_model_link_case <- function(data, model, link) {
  plot_data_one <- data |>
    filter(model_type == model, link_function == link)
  plot_summary_one <- publication_link_summaries |>
    filter(model_type == model, link_function == link)

  plot <- ggplot(plot_data_one, aes(CP_CV, community_stability, colour = experiment_label)) +
    geom_point(alpha = 0.32, size = 1.4) +
    geom_text(
      data = plot_summary_one,
      aes(x = Inf, y = Inf, label = panel_label),
      inherit.aes = FALSE,
      hjust = 1.08,
      vjust = 1.15,
      size = 3.7
    ) +
    scale_colour_manual(
      values = c("#3B5C7A", "#56B4E9", "#B56B45", "#009E73"),
      name = "Run"
    ) +
    labs(
      title = paste(model, "-", link),
      x = expression(CV(CP)),
      y = "Community stability, 1 / CV(total abundance)"
    ) +
    theme_figure(12)

  if (isTRUE(plot_summary_one$cp_cv_sd > 0 && is.finite(plot_summary_one$r_squared))) {
    plot <- plot + geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.75)
  }

  plot
}

plot_model_link_case_filtered <- function(data, model, link, max_cp_cv) {
  plot_data_one <- data |>
    filter(model_type == model, link_function == link, CP_CV <= max_cp_cv)

  plot_summary_one <- plot_data_one |>
    summarise(
      n = n(),
      cp_cv_sd = sd(CP_CV),
      stability_sd = sd(community_stability),
      r_squared = ifelse(
        n > 2 && cp_cv_sd > 0 && stability_sd > 0,
        summary(lm(community_stability ~ CP_CV))$r.squared,
        NA_real_
      ),
      .groups = "drop"
    ) |>
    mutate(
      panel_label = ifelse(
        is.na(r_squared),
        paste0("n = ", n, "\nCV(CP) constant"),
        paste0("n = ", n, "\nR² = ", formatC(r_squared, format = "f", digits = 2))
      )
    )

  plot <- ggplot(plot_data_one, aes(CP_CV, community_stability, colour = experiment_label)) +
    geom_point(alpha = 0.32, size = 1.4) +
    geom_text(
      data = plot_summary_one,
      aes(x = Inf, y = Inf, label = panel_label),
      inherit.aes = FALSE,
      hjust = 1.08,
      vjust = 1.15,
      size = 3.7
    ) +
    scale_colour_manual(
      values = c("#3B5C7A", "#56B4E9", "#B56B45", "#009E73"),
      name = "Run"
    ) +
    labs(
      title = paste0(model, " - ", link, " (CV(CP) <= ", max_cp_cv, ")"),
      x = expression(CV(CP)),
      y = "Community stability, 1 / CV(total abundance)"
    ) +
    theme_figure(12)

  if (isTRUE(plot_summary_one$cp_cv_sd > 0 && is.finite(plot_summary_one$r_squared))) {
    plot <- plot + geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.75)
  }

  plot
}

cr_identity_filter_threshold <- 50
```

### General Definition

A community performance curve ($CP$) can be defined by first mapping each
species' temperature-dependent performance onto the scale of the community
variable whose stability is being analysed, and then summing across species.

For stability of total abundance or total biomass, the species-level curve
therefore represents expected contribution to abundance or biomass, not
just raw physiological performance or intrinsic growth rate.

The general construction is:

1. Compute each species' temperature-dependent performance curve.
2. Transform each species' performance to the scale of the focal community
   variable.
3. Sum the transformed species curves to obtain the $CP$.
4. Use the temporal variability of that $CP$ to predict the stability of the
   focal community variable.

In notation:

$$
P_i(T) = \text{raw performance of species } i \text{ at temperature } T
$$

$$
A_i(T) = f_{\text{model}, y}(P_i(T))
$$

$$
CP_y(T) = \sum_i A_i(T)
$$

where $y$ is the community variable whose stability is being predicted, such as
total abundance or total biomass. The function $f_{\text{model}, y}$ is a
model-specific link from raw species performance to expected contribution to
that community variable.

### Why The Link Function Matters

When predicting the stability of total abundance or total biomass, the relevant
question is not simply whether species have high or low raw performance. The
relevant question is how changes in species performance translate into changes
in species abundance or biomass.

If raw performance is approximately linearly related to abundance, then summing
raw performance curves can be appropriate. But if abundance responds
nonlinearly, or if performance mainly determines whether a species can persist
at all, then raw performance is not on the scale of the response variable. In
that case, the species performance curve is first transformed onto an abundance
or biomass scale
before constructing the $CP$.

This gives the following definition:

> The $CP$ is a sum of species' expected contributions to the focal
> community variable, not necessarily a sum of raw species performance curves.

### Lotka-Volterra Model

In the Lotka-Volterra model used here, raw temperature-dependent performance is
closely linked to expected abundance by the model structure. Specifically, the
model links intrinsic growth rate and carrying capacity in a way that makes
higher performance correspond approximately linearly to higher expected
abundance.

For predicting stability of total abundance, the link function
from performance to abundance-scale contribution can be treated as approximately
identity:

$$
f_{\text{LV}, abundance}(P_i(T)) \approx P_i(T)
$$

So the LV abundance-scale $CP$ can be written as:

$$
CP_{\text{LV}}(T) = \sum_i P_i(T)
$$

Each species contributes its raw performance value to the community performance
curve. In the Lotka--Volterra simulations, the scale of the summed curve is
therefore close to the scale of the abundance response.

### Consumer-Resource Model

In the consumer-resource model, raw species performance is not linearly related
to equilibrium abundance. Instead, intrinsic growth rate mainly determines
whether a species can persist under the current thermal and resource context.

If growth is negative, the species declines to very low abundance. If growth is
positive, the species can grow, but its eventual abundance is then determined by
other conditions, including:

- resource supply;
- mortality;
- conversion efficiency;
- half-saturation constants;
- competition with other consumers for shared resources.

For predicting stability of total abundance, the abundance-scale transformation
is therefore represented by a viability threshold:

$$
f_{\text{CR}, abundance}(P_i(T)) =
\begin{cases}
1, & P_i(T) > 0 \\
0, & P_i(T) \le 0
\end{cases}
$$

The consumer-resource abundance-scale $CP$ is then:

$$
CP_{\text{CR viability}}(T) =
\sum_i I(P_i(T) > 0)
$$

where $I(P_i(T) > 0)$ is 1 when species $i$ has positive growth at temperature
$T$, and 0 otherwise.

The $CP$ is the number of viable species through time. Its $CV$ measures temporal
variation in the viable species pool under the temperature sequence.

### Raw and Viability $CP$

Raw $CV(CP)$ sums species growth-rate or performance curves. This construction is
appropriate for the Lotka--Volterra model because raw performance is
approximately on the abundance scale. In the consumer-resource model, raw growth
rate is not the abundance response. It primarily controls whether a species can
persist, while the abundance of persistent species is shaped by
resource-mediated interactions.

The viability-transformed $CP$ applies a model-specific link function before
summing:

$$
\text{raw performance} \rightarrow \text{expected abundance-scale contribution}
\rightarrow \text{community performance curve}
$$

### Proposed General Definition

For any model and any focal community variable, define a $CP$ as:

$$
CP_y(T) = \sum_i f_{\text{model}, y}(P_i(T))
$$

where:

- $P_i(T)$ is the raw species performance curve;
- $y$ is the community variable whose stability is being predicted;
- $f_{\text{model}, y}$ maps raw species performance to expected contribution
  to $y$;
- the $CP$ is formed only after this transformation.

For the current models and response variable:

```{r}
#| label: tbl-cp-link-definition
#| tbl-cap: "Model-specific $CP$ constructions for total-abundance stability."
#| tbl-cap-location: bottom

link_definition <- tibble(
  Model = c("Lotka-Volterra", "Consumer-resource"),
  `Focal stability variable` = c("Total abundance", "Total abundance"),
  `Link from species performance to contribution` = c(
    "Approximately identity",
    "Binary viability, I(P_i(T) > 0)"
  )
)

kable(
  link_definition
)
```

This definition keeps the $CP$ concept general while allowing the species-level
performance-to-abundance link to differ among models.

### Model-Link Cases

The figure panels below apply this construction to the simulation outputs.
Stability is shown as $1 / CV_{\text{total abundance}}$. The x-axis is $CV(CP)$
constructed under each model-link combination. The main simulation analyses in
@fig-main-simulation-results use the inverse response, total-abundance
instability.

For the Lotka-Volterra model, the identity-link $CP$ comes directly from the
community-measures output. The binary-link $CP$ uses the 
viability $CP$ calculated for the same runs. For the consumer-resource
model, both identity-link and binary-link $CV(CP)$ values come from the 
community-measures output.

#### Lotka-Volterra, Identity Link

```{r}
#| label: fig-lv-identity-link
#| fig-cap: "Lotka-Volterra identity-link $CP$. Points show simulation cases, the x-axis shows $CV(CP)$ based on raw community performance, and the y-axis shows community stability as $1 / CV(\\text{total abundance})$. The identity link is the model-consistent $CP$ construction for the Lotka-Volterra simulations."
#| fig-width: 7
#| fig-height: 5
#| fig-cap-location: bottom

plot_model_link_case(publication_link_data, "Lotka--Volterra", "Identity link")
```

#### Lotka-Volterra, Binary Link

```{r}
#| label: fig-lv-binary-link
#| fig-cap: "Lotka-Volterra binary-link $CP$. Species performance is transformed into binary viability before constructing $CP$. This comparison evaluates the effect of reducing the Lotka-Volterra abundance response to a viability count rather than using the raw performance scale."
#| fig-width: 7
#| fig-height: 5
#| fig-cap-location: bottom

plot_model_link_case(publication_link_data, "Lotka--Volterra", "Binary link")
```

The binary-link $CP$ evaluates the effect of reducing the Lotka--Volterra response
to a viability count rather than using the raw performance scale that enters the
model dynamics directly.

#### Consumer-Resource, Identity Link

```{r}
#| label: fig-cr-identity-link
#| fig-cap: "Consumer-resource identity-link $CP$. Raw performance values are summed directly for the consumer-resource simulations, without first mapping performance onto an abundance-relevant viability scale."
#| fig-width: 7
#| fig-height: 5
#| fig-cap-location: bottom

plot_model_link_case(publication_link_data, "Consumer-resource", "Identity link")
```

The identity-link $CP$ in the consumer-resource model sums raw uptake or raw
performance directly rather than first mapping performance onto an
abundance-relevant viability scale.

The same relationship with $CV(CP)$ values above `r cr_identity_filter_threshold`
excluded is:

```{r}
#| label: fig-cr-identity-link-filtered
#| fig-cap: "Consumer-resource identity-link $CP$ after excluding extreme raw $CV(CP)$ values. The filter removes cases with $CV(CP)$ above the stated threshold to show the relationship over the visually interpretable range of the raw identity-link metric."
#| fig-width: 7
#| fig-height: 5
#| fig-cap-location: bottom

plot_model_link_case_filtered(
  publication_link_data,
  "Consumer-resource",
  "Identity link",
  cr_identity_filter_threshold
)
```

#### Consumer-Resource, Binary Link

```{r}
#| label: fig-cr-binary-link
#| fig-cap: "Consumer-resource binary-link $CP$. Species performance is transformed into binary viability before constructing $CP$, matching the interpretation that total-abundance stability in the consumer-resource simulations is linked to the stability of the viable species pool."
#| fig-width: 7
#| fig-height: 5
#| fig-cap-location: bottom

plot_model_link_case(publication_link_data, "Consumer-resource", "Binary link")
```

The identity link was most closely aligned with the Lotka--Volterra model,
whereas the binary viability link was most closely aligned with the
consumer-resource model. The alternative link functions show how predictive
performance changes when $CP$ is constructed on a scale less directly tied to the
abundance variable whose stability is being predicted.

### Practical Interpretation

When the focal response is total abundance or biomass stability, $CP$ addresses
the following question:

> How much total abundance or biomass would the community be expected to support
> at each temperature, given the species that can contribute under that model?

In the LV model, species' raw performance values already approximate their
expected abundance contributions. In the consumer-resource model, the
first-order abundance contribution is closer to whether each species is viable
at all.

The predictive performance of viability $CP$ variability, $CV(CP)$, in the consumer-resource
experiments supports the interpretation that $CP$ is most informative when
constructed on the scale of the community variable whose stability is being
predicted.

### Summary of Raw and Viability $CP$

@fig-raw-vs-binary summarises the same link-function comparison as explanatory
power for actual-temperature $CP$ constructions, using the total-abundance
instability response used in the main simulation analyses.

```{r}
#| label: fig-raw-vs-binary
#| fig-cap: "Comparison of raw and binary $CP$ constructions for the Lotka-Volterra and consumer-resource simulations. Each point is the explanatory power of a single-predictor model relating total-abundance instability to one actual-temperature $CP$ variability metric, fitted separately for each run. Raw $CP$ sums expected performance or growth-rate values directly. Binary viability $CP$ counts species with positive expected net growth."
#| fig-width: 13
#| fig-height: 8
#| fig-cap-location: bottom

link_data <- analysis_data |>
  transmute(
    experiment,
    experiment_label,
    framework,
    framework_label,
    treatment_plot,
    stability,
    `Raw CP, actual` = raw_cpc,
    `Binary viability CP, actual` = viability_binary_cpc
  ) |>
  pivot_longer(
    cols = -c(experiment, experiment_label, framework, framework_label, treatment_plot, stability),
    names_to = "cp_metric",
    values_to = "cp_cv"
  ) |>
  filter(is.finite(stability), is.finite(cp_cv)) |>
  mutate(
    link = case_when(
      grepl("Raw", cp_metric) ~ "Raw CP",
      grepl("Binary", cp_metric) ~ "Binary viability CP"
    ),
    link = factor(link, levels = c("Raw CP", "Binary viability CP"))
  )

link_r2 <- link_data |>
  group_by(experiment, experiment_label, framework, framework_label, link) |>
  group_modify(~ tibble(
    n = nrow(.x),
    r2 = safe_r2(stability ~ cp_cv, .x)
  )) |>
  ungroup()

link_summary <- link_r2 |>
  group_by(link) |>
  summarise(
    median_r2 = median(r2, na.rm = TRUE),
    .groups = "drop"
  )

link_plot <- link_r2 |>
  ggplot(aes(x = link, y = r2, linetype = experiment_label, shape = experiment_label, group = experiment_label)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.8) +
  facet_wrap(~ framework_label, nrow = 1) +
  scale_linetype_manual(values = c("solid", "22", "42", "13"), name = "Run") +
  scale_shape_manual(values = c(16, 17, 15, 18), name = "Run") +
  scale_x_discrete(
    labels = c(
      "Raw CP" = expression(paste("Raw ", CP)),
      "Binary viability CP" = expression(paste("Binary viability ", CP))
    )
  ) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = NULL,
    y = expression(paste("Single-predictor explanatory power (", R^2, ")"))
  ) +
  theme_figure(12) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    strip.text.y = element_text(angle = 0)
  )

link_plot
```

The comparison supports the link-function interpretation. Actual binary and raw
$CP$ metrics differ in their explanatory power across modelling frameworks. In the
consumer-resource simulations, the binary viability scale corresponds more
closely to the abundance dynamics generated by resource-mediated competition.

## Additional Simulation Analyses

The supplementary analysis tests whether species thermal performance curves
predict the synchrony realised in the dynamical simulations, and whether this
relationship depends on interaction strength or resource sharing.

### Predicted Versus Realised Synchrony

Synchrony was calculated with the Loreau--de Mazancourt metric. For a set of
species-level time series $X_i(t)$, synchrony was

$$
\phi = \frac{\sigma^2\left(\sum_i X_i(t)\right)}
{\left(\sum_i \sigma(X_i(t))\right)^2}.
$$

For realised synchrony, $X_i(t)$ was the simulated abundance of species $i$.
For TPC-predicted synchrony, $X_i(t)$ was the species performance expected
under either the realised temperature sequence or the null temperature
distribution. Values near 0 indicate asynchronous species-level fluctuations,
whereas values near 1 indicate synchronous fluctuations.

```{r}
#| label: fig-predicted-realised-synchrony
#| fig-cap: "Comparison between synchrony predicted from species thermal performance curves and synchrony realised in the abundance simulations. Predicted synchrony was calculated from species TPCs under either the realised temperature series or the null temperature distribution. Realised synchrony was calculated from simulated species abundances. The top panels show the overall relationship; the dashed line is the 1:1 line and the black line is a linear fit. The bottom panels show the same relationship as R² values calculated separately for each interaction-strength or shared-resource-use level."
#| fig-width: 12
#| fig-height: 11
#| fig-cap-location: bottom

synchrony_data <- analysis_data |>
  transmute(
    experiment,
    experiment_label,
    framework,
    framework_label,
    treatment_plot,
    realised_synchrony = as.numeric(synchrony_abundance),
    `Actual TPC synchrony` = as.numeric(synchrony_performance_info),
    `Null TPC synchrony` = as.numeric(synchrony_performance_naive)
  ) |>
  pivot_longer(
    cols = c(`Actual TPC synchrony`, `Null TPC synchrony`),
    names_to = "synchrony_predictor",
    values_to = "predicted_synchrony"
  ) |>
  filter(is.finite(realised_synchrony), is.finite(predicted_synchrony)) |>
  mutate(
    synchrony_predictor = factor(
      synchrony_predictor,
      levels = c("Actual TPC synchrony", "Null TPC synchrony")
    )
  )

synchrony_summary <- synchrony_data |>
  group_by(framework, framework_label, synchrony_predictor) |>
  summarise(
    n = n(),
    r2 = summary(lm(realised_synchrony ~ predicted_synchrony))$r.squared,
    correlation = cor(realised_synchrony, predicted_synchrony),
    .groups = "drop"
  )

safe_sync_r2 <- function(data) {
  if (
    nrow(data) < 4 ||
      sd(data$realised_synchrony, na.rm = TRUE) == 0 ||
      sd(data$predicted_synchrony, na.rm = TRUE) == 0
  ) {
    return(NA_real_)
  }
  summary(lm(realised_synchrony ~ predicted_synchrony, data = data))$r.squared
}

synchrony_treatment_summary <- synchrony_data |>
  group_by(framework, framework_label, treatment_plot, synchrony_predictor) |>
  group_modify(~ tibble(
    n = nrow(.x),
    r2 = safe_sync_r2(.x),
    correlation = cor(.x$realised_synchrony, .x$predicted_synchrony)
  )) |>
  ungroup()

synchrony_treatment_table <- synchrony_treatment_summary |>
  mutate(r2 = round(r2, 3)) |>
  select(framework, treatment = treatment_plot, synchrony_predictor, r2) |>
  pivot_wider(names_from = synchrony_predictor, values_from = r2) |>
  arrange(framework, treatment)

get_sync_value <- function(framework_name, predictor_name, column = "r2") {
  synchrony_summary |>
    filter(framework == framework_name, synchrony_predictor == predictor_name) |>
    pull({{ column }}) |>
    first()
}

synchrony_plot <- synchrony_data |>
  ggplot(aes(x = predicted_synchrony, y = realised_synchrony, colour = treatment_plot)) +
  geom_point(alpha = 0.16, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey45") +
  geom_smooth(method = "lm", se = FALSE, colour = "black", linewidth = 0.75) +
  geom_text(
    data = synchrony_summary |>
      mutate(label = paste0("R² = ", round(r2, 2), "\nr = ", round(correlation, 2))),
    aes(x = 0.03, y = 0.97, label = label),
    inherit.aes = FALSE,
    hjust = 0,
    vjust = 1,
    size = 3.5
  ) +
  facet_grid(framework_label ~ synchrony_predictor) +
  scale_colour_viridis_d(option = "C", end = 0.9, name = "Treatment") +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = expression(atop("Performance synchrony", phi[perf])),
    y = expression(atop("Biomass synchrony", phi[biom]))
  ) +
  theme_figure(12) +
  theme(strip.text.y = element_text(angle = 0))

synchrony_r2_plot <- synchrony_treatment_summary |>
  ggplot(aes(x = treatment_plot, y = r2, colour = synchrony_predictor, group = synchrony_predictor)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~ framework_label, nrow = 1) +
  scale_colour_manual(values = c("#3B5C7A", "#B56B45"), name = NULL) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Interaction strength / shared-resource use",
    y = expression(paste("Synchrony relationship ", R^2))
  ) +
  theme_figure(12)

synchrony_plot / synchrony_r2_plot +
  plot_layout(heights = c(1.45, 1)) +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")",
    theme = theme(plot.tag = element_text(size = 16, face = "bold"))
  )
```

```{r}
#| label: tbl-synchrony-r2-by-treatment
#| tbl-cap: "Synchrony relationship R² by interaction strength or shared-resource use."
#| tbl-cap-location: bottom

kable(
  synchrony_treatment_table
)
```

The synchrony comparison differed between the two modelling frameworks. In the
Lotka--Volterra simulations, TPC-predicted synchrony had a weak relationship
with realised abundance synchrony
(\(R^2 = `r round(get_sync_value("Lotka-Volterra", "Actual TPC synchrony"), 2)`\)
for actual TPC synchrony and
\(R^2 = `r round(get_sync_value("Lotka-Volterra", "Null TPC synchrony"), 2)`\)
for null TPC synchrony). In the consumer-resource simulations, the relationship
had higher explanatory power
(\(R^2 = `r round(get_sync_value("Consumer-resource", "Actual TPC synchrony"), 2)`\)
for actual TPC synchrony and
\(R^2 = `r round(get_sync_value("Consumer-resource", "Null TPC synchrony"), 2)`\)
for null TPC synchrony). The consumer-resource model therefore carried more of
the performance-curve synchrony signal into realised abundance
synchrony. The treatment-specific \(R^2\) values show that this is not just an
artefact of pooling across the resource-sharing gradient. Lotka--Volterra
abundance synchrony was most predictable from TPC synchrony when interactions
were absent, and this relationship declined as interaction strength increased.
In the consumer-resource simulations, synchrony prediction had the highest
explanatory power when resource use was mostly private and declined as
shared-resource use increased. In both frameworks, interactions mediated how
much species-level thermal synchrony was retained in realised abundance
synchrony.
