---
title: "Manuscript figures"
subtitle: "Reproducible figure file for the CPC manuscript"
editor: source
execute:
  echo: true
  warning: false
  message: false
  cache: false
format:
  html:
    page-layout: full
    html-math-method: mathml
    toc: true
    toc-location: left
    toc-depth: 2
    embed-resources: true
    code-fold: true
    code-summary: "Show code"
---

```{r setup}
library(dplyr)
library(forcats)
library(ggalluvial)
library(ggdist)
library(ggplot2)
library(ggrepel)
library(knitr)
library(mgcv)
library(patchwork)
library(purrr)
library(ranger)
library(readr)
library(readxl)
library(stringr)
library(tibble)
library(tidyr)
library(viridis)
library(viridisLite)

repo_root <- "/Users/francesco/Documents/GitHub/CPCs_predict_CommunityStability"
empirical_root <- file.path(repo_root, "new_empirical_study")
publication_root <- file.path(repo_root, "publication_experiments", "data")
literature_file_path <- file.path(repo_root, "literature_review", "data", "literature_file.xls")

theme_manuscript <- function(base_size = 12) {
  theme_bw(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(colour = "grey92"),
      strip.text = element_text(face = "bold", colour = "black"),
      legend.position = "bottom",
      plot.tag = element_text(face = "bold")
    )
}

theme_clean <- function(base_size = 20) {
  theme_minimal(base_size = base_size) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold", colour = "black"),
      legend.position = "bottom",
      plot.tag = element_text(face = "bold")
    )
}

format_p <- function(p) {
  ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
}
```

```{r empirical-data-preparation, include=FALSE}
ts_df_raw <- read.csv(file.path(empirical_root, "data", "combined_timeseries_df.csv"))
TPC_timeseries_raw <- read.csv(file.path(empirical_root, "data", "TPC_timeseries_df.csv"))

TPC_timeseries <- TPC_timeseries_raw

TPC_dens_df <- TPC_timeseries |>
  filter(species != "S") |>
  mutate(
    log_dens = log10(mean.dens.ml),
    max_days = case_when(
      species == "C" ~ 3,
      species == "D" & temperature %in% c("T24", "T27", "T33") ~ 2,
      species == "D" & temperature %in% c("T30") ~ 1,
      species == "D" ~ 3,
      species == "L" & temperature %in% c("T18", "T21", "T24", "T27", "T30", "T33") ~ 2,
      species == "L" ~ 3,
      species == "P" & temperature %in% c("T24", "T27") ~ 2,
      species == "P" & temperature %in% c("T15", "T18") ~ 6,
      species == "P" & temperature %in% c("T30", "T33") ~ 2,
      species == "P" ~ 3,
      species == "T" & temperature %in% c("T30", "T33") ~ 1,
      species == "T" & temperature == "T27" ~ 2,
      species == "T" & temperature %in% c("T18", "T21", "T24") ~ 3,
      species == "T" & temperature == "T15" ~ 4
    )
  ) |>
  filter(day <= max_days) |>
  group_by(species, temperature) |>
  summarise(
    igr = coef(lm(log_dens ~ day))[2],
    n_days = n_distinct(day),
    .groups = "drop"
  ) |>
  mutate(temperature = as.numeric(sub("T", "", temperature)))

new_data <- data.frame(temperature = seq(15, 33, by = 0.02))

nested_gams <- TPC_dens_df |>
  nest(cols = -species) |>
  mutate(
    gams = map(
      cols,
      ~ gam(igr ~ s(temperature, k = 7), data = .x, method = "REML")
    ),
    predicted = map(gams, ~ predict(.x, newdata = new_data))
  )

tpc_gams_design <- setNames(nested_gams$gams, nested_gams$species)

TPCs <- nested_gams |>
  unnest(predicted) |>
  bind_cols(new_data[rep(seq_len(nrow(new_data)), times = nrow(nested_gams)), , drop = FALSE]) |>
  select(species, temperature, predicted)

species_list <- c("L", "D", "P", "C", "T")
richness_levels <- c(2, 3, 4)

all_combinations <- map_dfr(richness_levels, function(richness_value) {
  combn(species_list, richness_value, simplify = FALSE) |>
    map_dfr(\(species_vector) {
      tibble(
        richness = richness_value,
        composition_name = paste(sort(species_vector), collapse = ""),
        species = list(species_vector)
      )
    })
})

list_combinations <- split(all_combinations, all_combinations$composition_name)

calculate_null_cv <- function() {
  lapply(list_combinations, function(row_data) {
    community <- row_data[["species"]][[1]]

    selected <- TPCs |>
      filter(species %in% community)

    sum_curve <- selected |>
      group_by(temperature) |>
      summarise(tot_igr = sum(predicted), .groups = "drop")

    pop_CV <- selected |>
      group_by(species) |>
      summarise(
        sd = sd(predicted),
        mean = mean(predicted),
        weight = mean / mean(sum_curve$tot_igr),
        CV = sd / mean,
        .groups = "drop"
      ) |>
      summarise(
        pop_CVperf_null = sum(CV * weight),
        synchrony_null = (sd(sum_curve$tot_igr))^2 / (sum(sd))^2
      )

    tibble(
      CV_perf_null = sd(sum_curve$tot_igr) / mean(sum_curve$tot_igr),
      pop_CVperf_null = pop_CV$pop_CVperf_null,
      synchrony_null = pop_CV$synchrony_null,
      composition_name = row_data$composition_name,
      richness = row_data$richness
    )
  }) |>
    bind_rows()
}

make_reddened_temperature <- function(seed) {
  n <- 35
  gamma <- 0.8
  target_mean <- 22.5
  target_sd <- 3.5
  temp_step <- 0.1

  set.seed(seed)
  f <- seq(1, n / 2)
  scales <- 1 / (f^(gamma / 2))
  phases <- runif(length(f), 0, 2 * pi)
  coeffs <- scales * exp(1i * phases)
  full_coeffs <- c(0, coeffs, rev(Conj(coeffs[-length(coeffs)])))
  raw_noise <- Re(fft(full_coeffs, inverse = TRUE))[1:n]
  temperature_reddened <- as.vector(scale(raw_noise)) * target_sd + target_mean
  temperature_reddened <- round(temperature_reddened / temp_step) * temp_step

  tibble(
    time = seq_len(n),
    temp = temperature_reddened
  ) |>
    drop_na()
}

df_red1 <- make_reddened_temperature(246)
df_red2 <- make_reddened_temperature(113)
df_red3 <- make_reddened_temperature(1356)
df_red4 <- make_reddened_temperature(3356)

df_temp <- tibble(
  time = df_red1$time,
  TS1 = df_red1$temp,
  TS2 = df_red2$temp,
  TS3 = df_red3$temp,
  TS4 = df_red4$temp
)

temp_series <- df_temp |>
  pivot_longer(cols = starts_with("TS"), values_to = "temp", names_to = "temp_series")

experimental_day_index <- c(1:5, 8:12, 15:19, 22:26, 29:33)

calculate_actual_cv <- function(T_norm) {
  lapply(list_combinations, function(row_data) {
    community <- row_data[["species"]][[1]]

    selected <- TPCs |>
      filter(species %in% community, temperature %in% T_norm)

    sum_curve <- selected |>
      group_by(temperature) |>
      summarise(tot_igr = sum(predicted), .groups = "drop")

    pop_CV <- selected |>
      group_by(species) |>
      summarise(
        sd = sd(predicted),
        mean = mean(predicted),
        weight = mean / mean(sum_curve$tot_igr),
        CV = sd / mean,
        .groups = "drop"
      ) |>
      summarise(
        pop_CVperf_actual = sum(CV * weight),
        synchrony_actual = (sd(sum_curve$tot_igr))^2 / (sum(sd))^2
      )

    tibble(
      CV_perf_actual = sd(sum_curve$tot_igr) / mean(sum_curve$tot_igr),
      pop_CVperf_actual = pop_CV$pop_CVperf_actual,
      synchrony_actual = pop_CV$synchrony_actual,
      composition_name = row_data$composition_name,
      richness = row_data$richness
    )
  }) |>
    bind_rows()
}

T_norm1 <- df_red1$temp[experimental_day_index]
T_norm2 <- df_red2$temp[experimental_day_index]
T_norm3 <- df_red3$temp[experimental_day_index]
T_norm4 <- df_red4$temp[experimental_day_index]

df_CVperf_naive_dens <- calculate_null_cv()
df_CVperf_obs_dens1 <- calculate_actual_cv(T_norm1)
df_CVperf_obs_dens2 <- calculate_actual_cv(T_norm2)
df_CVperf_obs_dens3 <- calculate_actual_cv(T_norm3)
df_CVperf_obs_dens4 <- calculate_actual_cv(T_norm4)

comp_choice1 <- df_CVperf_obs_dens1 |>
  filter(composition_name %in% c("PT", "CD", "LT", "CLT", "LPT", "DPT", "CDLT", "DLPT", "CDLP")) |>
  mutate(temp_series = 1)

comp_choice2 <- df_CVperf_obs_dens2 |>
  filter(composition_name %in% c("LT", "CP", "PT", "CLT", "DLP", "DPT", "CDLT", "CDLP", "CDPT")) |>
  mutate(temp_series = 2)

comp_choice3 <- df_CVperf_obs_dens3 |>
  filter(composition_name %in% c("LT", "CD", "PT", "CLT", "DLP", "DPT", "CDLT", "CDLP", "CDPT")) |>
  mutate(temp_series = 3)

comp_choice4 <- df_CVperf_obs_dens4 |>
  filter(composition_name %in% c("LT", "LP", "PT", "CLT", "LPT", "DPT", "CDLT", "CDLP", "CDPT")) |>
  mutate(temp_series = 4)

all_comps <- bind_rows(comp_choice1, comp_choice2, comp_choice3, comp_choice4)

species_T_opt <- TPCs |>
  filter(species != "S") |>
  group_by(species) |>
  slice_max(predicted, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(species, T_opt = temperature)

composition_T_opt_summary <- all_comps |>
  distinct(composition_name, richness) |>
  left_join(all_combinations, by = c("composition_name", "richness")) |>
  unnest(species) |>
  left_join(species_T_opt, by = "species") |>
  group_by(composition_name, richness) |>
  summarise(
    T_opt_mean = abs(mean(T_opt) - 22.5),
    sd_T_opt = sd(T_opt),
    .groups = "drop"
  )

CPCs <- all_comps |>
  left_join(df_CVperf_naive_dens, by = c("composition_name", "richness")) |>
  left_join(composition_T_opt_summary, by = c("composition_name", "richness")) |>
  mutate(
    temp_series = paste0("TS", temp_series),
    richness = as.factor(richness)
  )

ts_df <- ts_df_raw |>
  filter(day > 4) |>
  mutate(
    day = day - 4,
    richness = as.factor(nchar(composition_name))
  )

outlier_sampling_days <- tribble(
  ~sample_ID, ~day,
  "DLP_TS3_2", 4,
  "PT_TS4_1", 1,
  "LP_TS4_1", 3
)

ts_df <- ts_df |>
  anti_join(outlier_sampling_days, by = c("sample_ID", "day"))

get_pooled_gam_trend <- function(y, day) {
  if (length(y) >= 4 && n_distinct(day, na.rm = TRUE) >= 4) {
    trend_data <- tibble(y = y, day = day)
    trend_model <- gam(y ~ s(day, k = 5), data = trend_data, method = "REML")
    predict(trend_model, newdata = trend_data)
  } else {
    rep(mean(y, na.rm = TRUE), length(y))
  }
}

ts_df <- ts_df |>
  group_by(composition_name, predict_spec) |>
  mutate(
    mean.dens.ml_trend = get_pooled_gam_trend(mean.dens.ml, day),
    biomass_trend = get_pooled_gam_trend(biomass, day),
    mean.dens.ml_detrended = mean.dens.ml - mean.dens.ml_trend + mean(mean.dens.ml_trend, na.rm = TRUE),
    biomass_detrended = biomass - biomass_trend
  ) |>
  ungroup()

CV_0_df <- ts_df |>
  group_by(sample_ID, composition_name, temp_series, replicate, richness, day) |>
  summarise(com_biom = sum(biomass_detrended), .groups = "drop") |>
  group_by(sample_ID, composition_name, temp_series, replicate, richness) |>
  summarise(
    mean_biom = mean(com_biom),
    sd_biom = sd(com_biom),
    .groups = "drop"
  )

pop_CV <- ts_df |>
  group_by(sample_ID, composition_name, temp_series, replicate, richness, predict_spec) |>
  summarise(sd_pop_biom = sd(biomass_detrended), .groups = "drop") |>
  left_join(
    CV_0_df,
    by = c("sample_ID", "composition_name", "temp_series", "replicate", "richness")
  ) |>
  group_by(sample_ID, composition_name, temp_series, replicate, richness) |>
  summarise(
    mean_sd_pop = mean(sd_pop_biom),
    synchrony = unique(sd_biom)^2 / (sum(sd_pop_biom))^2,
    .groups = "drop"
  )

CV_0_df <- CV_0_df |>
  left_join(
    pop_CV,
    by = c("sample_ID", "composition_name", "temp_series", "replicate", "richness")
  )

CPC_CV_df <- CV_0_df |>
  left_join(
    CPCs,
    by = c("composition_name", "temp_series", "richness")
  ) |>
  mutate(richness_num = as.numeric(as.character(richness)))

m_CPC_obs <- lm(data = CPC_CV_df, log10(sd_biom) ~ log10(CV_perf_actual))
m_CPC_naive <- lm(data = CPC_CV_df, log10(sd_biom) ~ log10(CV_perf_null))
m_T_opt_mean <- lm(data = CPC_CV_df, log10(sd_biom) ~ T_opt_mean)
m_T_opt_sd <- lm(data = CPC_CV_df, log10(sd_biom) ~ sd_T_opt)

r2_obs <- summary(m_CPC_obs)$r.squared
r2_naive <- summary(m_CPC_naive)$r.squared
r2_T_opt_mean <- summary(m_T_opt_mean)$r.squared
r2_T_opt_sd <- summary(m_T_opt_sd)$r.squared

m_pop_var <- lm(data = CPC_CV_df, mean_sd_pop ~ pop_CVperf_actual)
sum_pop_var <- summary(m_pop_var)
pop_var_beta_text <- formatC(coef(sum_pop_var)["pop_CVperf_actual", "Estimate"], format = "e", digits = 2)
pop_var_p_text <- format_p(coef(sum_pop_var)["pop_CVperf_actual", "Pr(>|t|)"])
pop_var_r2_text <- sprintf("%.2f", sum_pop_var$r.squared)

m_syn <- lm(data = CPC_CV_df, synchrony ~ synchrony_actual)
sum_syn <- summary(m_syn)
syn_beta_text <- formatC(coef(sum_syn)["synchrony_actual", "Estimate"], format = "e", digits = 2)
syn_p_text <- format_p(coef(sum_syn)["synchrony_actual", "Pr(>|t|)"])
syn_r2_text <- sprintf("%.2f", sum_syn$r.squared)
```

# Figure 1 - empirical version

Although this figure is NOT included in the manuscript, it shows how one could potentially recreate a very similar figure from empirical data (in this case, from out ciliate microcosm experiment). Code to reproduce the figure shown in the manuscript is right below. 
```{r figure-1, fig.width=15, fig.height=12}
figure1_communities <- c("PT", "LT", "DPT", "CDLP")
figure1_actual_temperatures <- tibble(
  time = seq_along(T_norm2),
  temperature = T_norm2
)
figure1_null_temperatures <- tibble(
  time = seq_along(T_norm2),
  temperature = seq(min(TPCs$temperature), max(TPCs$temperature), length.out = length(T_norm2))
)

species_colours <- setNames(
  viridisLite::magma(length(species_list), begin = 0.20, end = 0.85),
  species_list
)

community_label <- function(composition_name) {
  paste(strsplit(composition_name, "")[[1]], collapse = " + ")
}

predict_species_performance <- function(composition_name, temperature_df) {
  community_species <- strsplit(composition_name, "")[[1]]

  crossing(
    temperature_df,
    species = community_species
  ) |>
    mutate(
      performance = map2_dbl(
        species,
        temperature,
        ~ as.numeric(predict(
          tpc_gams_design[[.x]],
          newdata = tibble(temperature = .y)
        ))
      )
    )
}

community_cp_timeseries <- function(composition_name, temperature_df) {
  predict_species_performance(composition_name, temperature_df) |>
    group_by(time) |>
    mutate(community_performance = sum(performance)) |>
    ungroup()
}

scale_temperature_trace <- function(temperature_df, y_values) {
  y_range <- range(y_values, na.rm = TRUE, finite = TRUE)
  temp_range <- range(temperature_df$temperature, na.rm = TRUE, finite = TRUE)

  if (!all(is.finite(y_range)) || diff(y_range) == 0 ||
      !all(is.finite(temp_range)) || diff(temp_range) == 0) {
    return(mutate(temperature_df, temperature_scaled = mean(y_range)))
  }

  temperature_df |>
    mutate(
      temperature_scaled = y_range[1] +
        (temperature - temp_range[1]) / diff(temp_range) * diff(y_range)
    )
}

curve_column <- function(composition_name, show_title = FALSE, show_x = FALSE) {
  community_species <- strsplit(composition_name, "")[[1]]

  species_curves <- TPCs |>
    filter(species %in% community_species)

  cp_curve <- species_curves |>
    group_by(temperature) |>
    summarise(community_performance = sum(predicted), .groups = "drop")

  p <- ggplot() +
    geom_line(
      data = species_curves,
      aes(x = temperature, y = predicted, colour = species),
      linewidth = 1.05,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_curve,
      aes(x = temperature, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    scale_colour_manual(
      values = species_colours,
      limits = species_list,
      breaks = species_list,
      drop = FALSE
    ) +
    theme_clean(12) +
    labs(
      title = if (show_title) "Species curves and CP" else NULL,
      subtitle = community_label(composition_name),
      x = "Temperature (°C)",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

timeseries_column <- function(composition_name, temperature_df, title, show_x = FALSE) {
  plot_data <- community_cp_timeseries(composition_name, temperature_df)
  cp_data <- plot_data |>
    distinct(time, community_performance)

  cp_cv <- sd(cp_data$community_performance) / mean(cp_data$community_performance)
  temperature_trace <- scale_temperature_trace(
    temperature_df,
    c(plot_data$performance, cp_data$community_performance)
  )

  p <- ggplot() +
    geom_line(
      data = temperature_trace,
      aes(x = time, y = temperature_scaled),
      inherit.aes = FALSE,
      colour = "grey35",
      linewidth = 0.7,
      linetype = "22",
      alpha = 0.45
    ) +
    geom_line(
      data = plot_data,
      aes(x = time, y = performance, colour = species, group = species),
      linewidth = 0.95,
      alpha = 0.78
    ) +
    geom_point(
      data = plot_data,
      aes(x = time, y = performance, colour = species),
      size = 1.4,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_data,
      aes(x = time, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    annotate(
      "label",
      x = Inf,
      y = Inf,
      hjust = 1.03,
      vjust = 1.12,
      label = paste0("CV(CP) = ", sprintf("%.2f", cp_cv)),
      label.size = 0.18,
      size = 3.4
    ) +
    scale_colour_manual(
      values = species_colours,
      limits = species_list,
      breaks = species_list,
      drop = FALSE
    ) +
    theme_clean(12) +
    labs(
      title = title,
      x = "Time step",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

figure1_rows <- imap(figure1_communities, function(composition_name, row_index) {
  show_title <- row_index == 1
  show_x <- row_index == length(figure1_communities)

  curve_column(composition_name, show_title = show_title, show_x = show_x) |
    timeseries_column(
      composition_name,
      figure1_null_temperatures,
      title = if (show_title) "Null CP over uniform temperatures" else NULL,
      show_x = show_x
    ) |
    timeseries_column(
      composition_name,
      figure1_actual_temperatures,
      title = if (show_title) "Actual CP over experienced temperatures" else NULL,
      show_x = show_x
    )
})

wrap_plots(figure1_rows, ncol = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "none")
```

Figure 1, empirical version. Conceptual construction of community performance (CP) variability using the empirical ciliate thermal-performance curves. The first column shows species-level thermal performance curves and the summed CP curve for four example communities. The second column evaluates the same species curves over a uniform temperature sequence to produce null CP time series and their CV. The third column evaluates the same curves over one realised experimental temperature sequence to produce actual CP time series and their CV. Thin coloured lines are species-level performance, thick black lines are summed community performance, and dashed grey lines in the time-series panels show the temperature sequence rescaled to the plotting height.

## Figure 1 - Main Text

```{r figure-1-conceptual, fig.width=15, fig.height=12}
concept_species_colours <- setNames(
  viridisLite::magma(5, begin = 0.20, end = 0.85),
  paste0("sp", 1:5)
)

concept_temperature_grid <- seq(10, 30, length.out = 250)

concept_null_temperatures <- tibble(
  time = seq_len(25),
  temperature = seq(10, 30, length.out = 25)
)

concept_actual_temperatures <- tibble(
  time = seq_along(T_norm2),
  temperature = T_norm2
)

concept_species_sets <- list(
  "Widely spaced optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(12, 15, 20, 25, 28),
    width = c(3, 5, 4, 6, 3)
  ),
  "Tightly clustered optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(19, 19.5, 20, 20.5, 21),
    width = c(3.5, 3.8, 4, 3.6, 3.9)
  ),
  "Cool-skewed optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(11, 12, 13, 14, 15),
    width = c(3, 4, 5, 3, 4)
  ),
  "Bimodal optima" = tibble(
    species = paste0("sp", 1:5),
    Topt = c(11, 12, 13, 27, 29),
    width = c(3, 4, 5, 3, 4)
  )
)

concept_performance <- function(temperature, optimum, width) {
  exp(-0.5 * ((temperature - optimum) / width)^2)
}

concept_species_performance <- function(species_df, temperature_df) {
  crossing(
    temperature_df,
    species_df
  ) |>
    mutate(
      performance = concept_performance(temperature, Topt, width)
    )
}

concept_cp_timeseries <- function(species_df, temperature_df) {
  concept_species_performance(species_df, temperature_df) |>
    group_by(time) |>
    mutate(community_performance = sum(performance)) |>
    ungroup()
}

concept_viability_timeseries <- function(species_df, temperature_df, threshold = 0.35) {
  concept_species_performance(species_df, temperature_df) |>
    mutate(viable = as.numeric(performance >= threshold)) |>
    group_by(time) |>
    mutate(community_viability = mean(viable)) |>
    ungroup()
}

concept_curve_column <- function(species_df, community_label, show_title = FALSE, show_x = FALSE) {
  curve_temperatures <- tibble(
    time = seq_along(concept_temperature_grid),
    temperature = concept_temperature_grid
  )

  species_curves <- concept_species_performance(species_df, curve_temperatures)

  cp_curve <- species_curves |>
    group_by(temperature) |>
    summarise(community_performance = sum(performance), .groups = "drop")

  p <- ggplot() +
    geom_line(
      data = species_curves,
      aes(x = temperature, y = performance, colour = species),
      linewidth = 1.05,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_curve,
      aes(x = temperature, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    theme_clean(12) +
    labs(
      title = if (show_title) "Species curves and CP" else NULL,
      subtitle = community_label,
      x = "Temperature (°C)",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_timeseries_column <- function(species_df, temperature_df, title, show_x = FALSE) {
  plot_data <- concept_cp_timeseries(species_df, temperature_df)
  cp_data <- plot_data |>
    distinct(time, community_performance)

  cp_cv <- sd(cp_data$community_performance) / mean(cp_data$community_performance)
  temperature_trace <- scale_temperature_trace(
    temperature_df,
    c(plot_data$performance, cp_data$community_performance)
  )

  p <- ggplot() +
    geom_line(
      data = temperature_trace,
      aes(x = time, y = temperature_scaled),
      inherit.aes = FALSE,
      colour = "grey35",
      linewidth = 0.7,
      linetype = "22",
      alpha = 0.45
    ) +
    geom_line(
      data = plot_data,
      aes(x = time, y = performance, colour = species, group = species),
      linewidth = 0.95,
      alpha = 0.78
    ) +
    geom_point(
      data = plot_data,
      aes(x = time, y = performance, colour = species),
      size = 1.4,
      alpha = 0.78
    ) +
    geom_line(
      data = cp_data,
      aes(x = time, y = community_performance),
      linewidth = 1.55,
      colour = "black"
    ) +
    annotate(
      "label",
      x = Inf,
      y = Inf,
      hjust = 1.03,
      vjust = 1.12,
      label = paste0("CV(CP) = ", sprintf("%.2f", cp_cv)),
      label.size = 0.18,
      size = 3.4
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    theme_clean(12) +
    labs(
      title = title,
      x = "Time step",
      y = "Performance",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_species_sets <- concept_species_sets[
  names(concept_species_sets) != "Cool-skewed optima"
]

concept_figure_rows <- imap(concept_species_sets, function(species_df, community_label) {
  row_index <- match(community_label, names(concept_species_sets))
  show_title <- row_index == 1
  show_x <- row_index == length(concept_species_sets)

  concept_curve_column(
    species_df,
    community_label,
    show_title = show_title,
    show_x = show_x
  ) |
    concept_timeseries_column(
      species_df,
      concept_null_temperatures,
      title = if (show_title) "Null CP over uniform temperatures" else NULL,
      show_x = show_x
    ) |
    concept_timeseries_column(
      species_df,
      concept_actual_temperatures,
      title = if (show_title) "Actual CP over experienced temperatures" else NULL,
      show_x = show_x
    )
})

wrap_plots(concept_figure_rows, ncol = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "none")
```

Figure 1, conceptual-community version. This alternative uses 3 idealised communities with widely spaced, tightly clustered, or bimodal species optima. The layout is the same as in the empirical version: species-level performance and summed CP curves are shown in the first column, null-uniform CP time series and CVs in the second column, and actual-environment CP time series and CVs in the third column. Dashed grey lines in the time-series panels show the temperature sequence rescaled to the plotting height.

## Supplementary Figure 1.1

This figure is similar to previous one (Fig.1 in the main text), but it also shows the viability curves. 

```{r figure-1-viability-supplement, fig.width=15, fig.height=12}
viability_threshold <- 0.35

concept_viability_curve_column <- function(species_df, community_label, show_title = FALSE, show_x = FALSE) {
  curve_temperatures <- tibble(
    time = seq_along(concept_temperature_grid),
    temperature = concept_temperature_grid
  )

  species_curves <- concept_species_performance(species_df, curve_temperatures)

  viability_curve <- species_curves |>
    mutate(viable = as.numeric(performance >= viability_threshold)) |>
    group_by(temperature) |>
    summarise(community_viability = mean(viable), .groups = "drop")

  p <- ggplot() +
    geom_line(
      data = species_curves,
      aes(x = temperature, y = performance, colour = species),
      linewidth = 1.05,
      alpha = 0.70
    ) +
    geom_hline(
      yintercept = viability_threshold,
      colour = "grey35",
      linewidth = 0.55,
      linetype = "22"
    ) +
    geom_step(
      data = viability_curve,
      aes(x = temperature, y = community_viability),
      linewidth = 1.55,
      colour = "black"
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    coord_cartesian(ylim = c(0, 1.05)) +
    theme_clean(12) +
    labs(
      title = if (show_title) "Species curves and viability CP" else NULL,
      subtitle = community_label,
      x = "Temperature (°C)",
      y = "Performance / fraction viable",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

concept_viability_timeseries_column <- function(species_df, temperature_df, title, show_x = FALSE) {
  plot_data <- concept_viability_timeseries(
    species_df,
    temperature_df,
    threshold = viability_threshold
  )

  viability_data <- plot_data |>
    distinct(time, community_viability)

  viability_cv <- sd(viability_data$community_viability) /
    mean(viability_data$community_viability)

  temperature_trace <- scale_temperature_trace(
    temperature_df,
    c(plot_data$viable, viability_data$community_viability)
  )

  p <- ggplot() +
    geom_line(
      data = temperature_trace,
      aes(x = time, y = temperature_scaled),
      inherit.aes = FALSE,
      colour = "grey35",
      linewidth = 0.7,
      linetype = "22",
      alpha = 0.45
    ) +
    geom_line(
      data = plot_data,
      aes(x = time, y = viable, colour = species, group = species),
      linewidth = 0.85,
      alpha = 0.65
    ) +
    geom_point(
      data = plot_data,
      aes(x = time, y = viable, colour = species),
      size = 1.4,
      alpha = 0.70
    ) +
    geom_step(
      data = viability_data,
      aes(x = time, y = community_viability),
      linewidth = 1.55,
      colour = "black"
    ) +
    annotate(
      "label",
      x = Inf,
      y = Inf,
      hjust = 1.03,
      vjust = 1.12,
      label = paste0("CV(CP) = ", sprintf("%.2f", viability_cv)),
      label.size = 0.18,
      size = 3.4
    ) +
    scale_colour_manual(values = concept_species_colours, drop = FALSE) +
    coord_cartesian(ylim = c(-0.04, 1.05)) +
    theme_clean(12) +
    labs(
      title = title,
      x = "Time step",
      y = "Viability / fraction viable",
      colour = "Species"
    )

  if (!show_x) {
    p <- p + theme(axis.title.x = element_blank())
  }

  p
}

viability_figure_rows <- imap(concept_species_sets, function(species_df, community_label) {
  row_index <- match(community_label, names(concept_species_sets))
  show_title <- row_index == 1
  show_x <- row_index == length(concept_species_sets)

  concept_viability_curve_column(
    species_df,
    community_label,
    show_title = show_title,
    show_x = show_x
  ) |
    concept_viability_timeseries_column(
      species_df,
      concept_null_temperatures,
      title = if (show_title) "Null viability CP\nuniform temperatures" else NULL,
      show_x = show_x
    ) |
    concept_viability_timeseries_column(
      species_df,
      concept_actual_temperatures,
      title = if (show_title) "Actual viability CP\nexperienced temperatures" else NULL,
      show_x = show_x
    )
})

wrap_plots(viability_figure_rows, ncol = 1) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(legend.position = "none")
```

Supplementary Figure 1.1. Conceptual construction of viability CP. Species performance curves are first transformed into binary viability using the dashed threshold. The community viability CP is the fraction of species above that threshold at each temperature or time step. Dashed grey traces in the time-series panels show the temperature sequence rescaled to the panel height; black curves show community-level viability CP.

<!-- # Figure 2 -->

<!-- ```{r figure-2, fig.width=14, fig.height=8} -->
<!-- literature_file <- read_excel(literature_file_path, sheet = "Sheet1") -->

<!-- df_inc <- literature_file |> -->
<!--   filter(`included after first screening` == "yes") |> -->
<!--   select( -->
<!--     included_after_first_screening = `included after first screening`, -->
<!--     included_after_reading = `included after reading`, -->
<!--     type_of_study = `type of study`, -->
<!--     species_specific_responses = `species specific responses useful?`, -->
<!--     interactions_useful = `interactions useful?`, -->
<!--     ecosystem, -->
<!--     organism_group = `organism group`, -->
<!--     prevailing_mechanism = `prevailing mechanism` -->
<!--   ) |> -->
<!--   mutate(across(everything(), ~ forcats::fct_na_value_to_level(as.factor(.x), "Unknown"))) |> -->
<!--   filter(included_after_reading == "yes") -->

<!-- shared_theme <- theme_minimal(base_size = 16, base_family = "sans") + -->
<!--   theme( -->
<!--     plot.title = element_text(size = 16, face = "bold"), -->
<!--     axis.title = element_text(size = 15), -->
<!--     axis.text = element_text(size = 13), -->
<!--     legend.title = element_text(size = 13), -->
<!--     legend.text = element_text(size = 12), -->
<!--     legend.position = "bottom" -->
<!--   ) -->

<!-- organism_cols <- viridis( -->
<!--   n = length(unique(df_inc$organism_group)), -->
<!--   option = "magma", -->
<!--   begin = 0.35, -->
<!--   end = 0.85 -->
<!-- ) -->

<!-- alluvial <- ggplot( -->
<!--   df_inc, -->
<!--   aes( -->
<!--     axis3 = type_of_study, -->
<!--     axis4 = species_specific_responses, -->
<!--     axis5 = interactions_useful, -->
<!--     y = 1 -->
<!--   ) -->
<!-- ) + -->
<!--   geom_alluvium(aes(fill = organism_group), alpha = 0.85) + -->
<!--   geom_stratum(color = "grey70") + -->
<!--   geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.4) + -->
<!--   scale_x_discrete( -->
<!--     limits = c("Study type", "Species-specific?", "Interactions?"), -->
<!--     labels = c( -->
<!--       "Type of study", -->
<!--       "Species-specific\nresponses useful?", -->
<!--       "Interactions\nuseful?" -->
<!--     ) -->
<!--   ) + -->
<!--   scale_fill_manual( -->
<!--     values = organism_cols, -->
<!--     guide = guide_legend(nrow = 2, byrow = TRUE) -->
<!--   ) + -->
<!--   labs( -->
<!--     x = NULL, -->
<!--     y = "Number of studies", -->
<!--     fill = "Organism group" -->
<!--   ) + -->
<!--   scale_y_continuous(limits = c(0, 12), breaks = seq(0, 12, 2)) + -->
<!--   shared_theme -->

<!-- df_yes_yes <- df_inc |> -->
<!--   filter( -->
<!--     species_specific_responses == "yes", -->
<!--     interactions_useful == "yes" -->
<!--   ) -->

<!-- study_cols <- viridis( -->
<!--   n = length(unique(df_yes_yes$type_of_study)), -->
<!--   option = "mako", -->
<!--   begin = 0.35, -->
<!--   end = 0.85 -->
<!-- ) -->

<!-- p_prev_mech_by_study <- ggplot( -->
<!--   df_yes_yes, -->
<!--   aes(x = prevailing_mechanism, fill = type_of_study) -->
<!-- ) + -->
<!--   geom_bar(position = "stack", alpha = 0.85) + -->
<!--   coord_flip() + -->
<!--   scale_fill_manual( -->
<!--     values = study_cols, -->
<!--     guide = guide_legend(nrow = 1, byrow = TRUE) -->
<!--   ) + -->
<!--   labs( -->
<!--     title = "Prevailing mechanisms", -->
<!--     x = "Prevailing mechanism", -->
<!--     y = "Number of studies", -->
<!--     fill = "Study type" -->
<!--   ) + -->
<!--   shared_theme -->

<!-- alluvial / p_prev_mech_by_study + -->
<!--   plot_layout(heights = c(2.8, 1), guides = "keep") + -->
<!--   plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") & -->
<!--   theme(legend.position = "bottom") -->
<!-- ``` -->

<!-- Figure 2. Overview of study characteristics and prevailing mechanisms in the literature review. The upper alluvial diagram shows study type, whether species-specific responses were considered useful, and whether interactions were considered useful. Colours indicate organism groups. The lower panel shows the prevailing mechanism among studies that considered both species-specific responses and interactions useful. -->



# Figure 2
```{r simulation-preparation, include=FALSE}
selected_experiments <- c("LV1", "LV2", "CR1", "CR2")

`%||%` <- function(x, y) {
  if (is.null(x) || length(x) == 0) y else x
}

format_experiment_label <- function(x) {
  recode(
    x,
    LV1 = "LV 1",
    LV2 = "LV 2",
    CR1 = "CR 1",
    CR2 = "CR 2"
  )
}

case_spec_value <- function(case_spec, path, default = NA) {
  parts <- strsplit(path, ".", fixed = TRUE)[[1]]
  value <- case_spec
  for (part in parts) {
    if (!is.list(value) || is.null(value[[part]])) {
      return(default)
    }
    value <- value[[part]]
  }
  if (length(value) != 1) {
    return(default)
  }
  value
}

case_spec_numeric <- function(data, path, default = NA_real_) {
  vapply(
    data$case_spec,
    function(spec) as.numeric(case_spec_value(spec, path, default)),
    numeric(1)
  )
}

case_spec_character <- function(data, path, default = NA_character_) {
  vapply(
    data$case_spec,
    function(spec) as.character(case_spec_value(spec, path, default)),
    character(1)
  )
}

nominal_lv_strength <- function(labels) {
  parsed <- sub(".*(?:mean|median)_([0-9.]+).*", "\\1", labels, perl = TRUE)
  parsed[parsed == labels] <- NA_character_
  as.numeric(parsed)
}

read_experiment <- function(experiment) {
  readRDS(file.path(publication_root, experiment, "community_measures.RDS")) |>
    as_tibble() |>
    mutate(
      experiment = experiment,
      framework = if_else(grepl("^LV", experiment), "Lotka-Volterra", "Consumer-resource")
    )
}

standardise_experiment <- function(experiment) {
  dat <- read_experiment(experiment)

  if (grepl("^LV", experiment)) {
    labels <- case_spec_character(dat, "interactions.selected.label")
    dat |>
      mutate(
        treatment_label = labels,
        treatment_value = nominal_lv_strength(labels),
        treatment_plot = as.character(round(treatment_value, 2)),
        treatment_axis = "Nominal interaction strength",
        design_mean_optimum = case_spec_numeric(dat, "traits.birth_optimum.mean"),
        design_optimum_range = case_spec_numeric(dat, "traits.birth_optimum.range"),
        design_width_mean = case_spec_numeric(dat, "traits.birth_width.mean"),
        design_width_range = case_spec_numeric(dat, "traits.birth_width.range"),
        actual_cpc = as.numeric(cv_community_performance_info),
        null_cpc = as.numeric(cv_community_performance_naive)
      )
  } else {
    private_use <- case_spec_numeric(dat, "resources.private_use.mean")
    dat |>
      mutate(
        private_use_mean = private_use,
        treatment_value = 1 - private_use_mean,
        treatment_value = case_when(
          treatment_value <= 0.025 ~ 0,
          treatment_value >= 0.975 ~ 1,
          TRUE ~ treatment_value
        ),
        treatment_plot = as.character(round(treatment_value, 2)),
        treatment_label = paste0("shared_resource_", treatment_plot),
        treatment_axis = "Shared-resource use",
        design_mean_optimum = case_spec_numeric(dat, "traits.uptake_optimum.mean"),
        design_optimum_range = case_spec_numeric(dat, "traits.uptake_optimum.range"),
        design_width_mean = case_spec_numeric(dat, "traits.uptake_width.mean"),
        design_width_range = case_spec_numeric(dat, "traits.uptake_width.range"),
        actual_cpc = as.numeric(cv_community_viability_binary_info),
        null_cpc = as.numeric(cv_community_viability_binary_naive)
      )
  }
}

raw_data <- map_dfr(selected_experiments, standardise_experiment)

analysis_data <- raw_data |>
  mutate(
    stability = as.numeric(cv_total_abundance),
    distance_mean_optimum = abs(20 - as.numeric(realized_mean_performance_optimum)),
    sd_optimum = as.numeric(realized_sd_performance_optimum),
    framework_label = recode(
      framework,
      "Lotka-Volterra" = "Lotka-Volterra\nLV 1 + LV 2",
      "Consumer-resource" = "Consumer-resource\nCR 1 + CR 2"
    ),
    framework_label = factor(
      framework_label,
      levels = c("Lotka-Volterra\nLV 1 + LV 2", "Consumer-resource\nCR 1 + CR 2")
    ),
    experiment_label = factor(
      format_experiment_label(experiment),
      levels = format_experiment_label(selected_experiments)
    ),
    treatment_plot = factor(treatment_plot, levels = c("0", "0.25", "0.5", "0.75", "1"))
  ) |>
  filter(
    is.finite(stability),
    is.finite(actual_cpc),
    is.finite(null_cpc),
    is.finite(distance_mean_optimum),
    is.finite(sd_optimum),
    is.finite(richness),
    is.finite(treatment_value),
    !is.na(env_series_id)
  )

safe_r2 <- function(formula, data) {
  fit <- tryCatch(lm(formula, data = data), error = function(e) NULL)
  if (is.null(fit)) {
    return(NA_real_)
  }
  r2 <- summary(fit)$r.squared
  ifelse(is.finite(r2), r2, NA_real_)
}

grouped_r2 <- analysis_data |>
  group_by(experiment, framework, framework_label, treatment_plot, treatment_value, env_series_id) |>
  group_modify(~ {
    dat <- .x
    if (nrow(dat) < 4) {
      return(tibble(
        predictor = c("Actual CV(CP)", "Null CV(CP)", "Trait summaries"),
        r2 = NA_real_,
        n_cases = nrow(dat),
        note = "fewer than four observations in group"
      ))
    }

    tibble(
      predictor = c("Actual CV(CP)", "Null CV(CP)", "Trait summaries"),
      r2 = c(
        safe_r2(stability ~ actual_cpc, dat),
        safe_r2(stability ~ null_cpc, dat),
        safe_r2(stability ~ distance_mean_optimum * sd_optimum, dat)
      ),
      n_cases = nrow(dat),
      note = NA_character_
    )
  }) |>
  ungroup() |>
  mutate(
    predictor = factor(predictor, levels = c("Actual CV(CP)", "Null CV(CP)", "Trait summaries")),
    treatment_plot = factor(treatment_plot, levels = c("0", "0.25", "0.5", "0.75", "1"))
  )

rf_variable_labels <- c(
  actual_cpc = "Actual CV(CP)",
  null_cpc = "Null CV(CP)",
  distance_mean_optimum = "|mean optimum - mean T|",
  sd_optimum = "SD optimum",
  richness = "Richness",
  treatment_numeric = "Interaction/resource sharing"
)

make_rf_data <- function(experiment_name) {
  analysis_data |>
    filter(experiment == experiment_name) |>
    mutate(treatment_numeric = as.numeric(treatment_value)) |>
    select(
      experiment,
      experiment_label,
      framework,
      stability,
      actual_cpc,
      null_cpc,
      distance_mean_optimum,
      sd_optimum,
      richness,
      treatment_numeric
    ) |>
    filter(
      if_all(
        c(stability, actual_cpc, null_cpc, distance_mean_optimum, sd_optimum, richness, treatment_numeric),
        is.finite
      )
    )
}

fit_rf_outputs <- function(experiment_name) {
  dat <- make_rf_data(experiment_name)

  set.seed(123)
  rf <- ranger(
    stability ~ actual_cpc + null_cpc + distance_mean_optimum +
      sd_optimum + richness + treatment_numeric,
    data = dat,
    importance = "permutation",
    num.trees = 1000,
    mtry = 3
  )

  importance <- enframe(rf$variable.importance, name = "variable", value = "importance") |>
    mutate(
      experiment = experiment_name,
      experiment_label = format_experiment_label(experiment_name),
      scaled_importance = importance / max(importance, na.rm = TRUE)
    )

  predictions <- dat |>
    transmute(
      experiment = experiment_name,
      experiment_label = first(experiment_label),
      framework = first(framework),
      observed = stability,
      predicted = rf$predictions,
      oob_r2 = rf$r.squared
    )

  list(importance = importance, predictions = predictions)
}

rf_outputs <- map(selected_experiments, fit_rf_outputs)

rf_importance <- map_dfr(rf_outputs, "importance") |>
  mutate(
    variable_label = recode(variable, !!!rf_variable_labels),
    variable_label = factor(variable_label, levels = unname(rf_variable_labels)),
    experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments))
  )

rf_prediction_data <- map_dfr(rf_outputs, "predictions") |>
  filter(is.finite(observed), is.finite(predicted)) |>
  mutate(
    experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments)),
    oob_label = paste0("OOB R² = ", round(oob_r2, 3))
  )

random_subset_r2 <- function(experiment_name) {
  dat <- make_rf_data(experiment_name)
  n_groups <- min(500L, max(1L, floor(nrow(dat) / 25L)))

  set.seed(100 + match(experiment_name, selected_experiments))
  dat |>
    mutate(random_group = sample(rep(seq_len(n_groups), length.out = n()))) |>
    group_by(random_group) |>
    summarise(
      n_pairs = sum(is.finite(stability) & is.finite(actual_cpc)),
      r2 = if (n_pairs >= 4) {
        cor(stability, actual_cpc, use = "complete.obs")^2
      } else {
        NA_real_
      },
      .groups = "drop"
    ) |>
    filter(is.finite(r2)) |>
    mutate(
      experiment = experiment_name,
      experiment_label = format_experiment_label(experiment_name)
    )
}

random_subset_data <- map_dfr(selected_experiments, random_subset_r2) |>
  mutate(experiment_label = factor(experiment_label, levels = format_experiment_label(selected_experiments)))
```



```{r figure-3, fig.width=15, fig.height=10.5}
importance_offsets <- tibble(
  experiment_label = factor(
    format_experiment_label(selected_experiments),
    levels = format_experiment_label(selected_experiments)
  ),
  importance_offset = c(-0.24, -0.08, 0.08, 0.24)
)

importance_data <- rf_importance |>
  filter(is.finite(scaled_importance)) |>
  left_join(importance_offsets, by = "experiment_label") |>
  mutate(
    variable_label = factor(variable_label, levels = rev(unname(rf_variable_labels))),
    variable_y = as.numeric(variable_label) + importance_offset
  )

experiment_levels <- unique(c(
  as.character(random_subset_data$experiment_label),
  as.character(importance_data$experiment_label),
  as.character(rf_prediction_data$experiment_label)
))

experiment_colours <- setNames(
  magma(n = length(experiment_levels), begin = 0.20, end = 0.85),
  experiment_levels
)

framework_levels <- c("LV", "CR")

framework_colours <- setNames(
  magma(n = length(framework_levels), begin = 0.20, end = 0.85),
  framework_levels
)

random_subset_data <- random_subset_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

importance_data <- importance_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

rf_prediction_data <- rf_prediction_data |>
  mutate(experiment_label = factor(experiment_label, levels = experiment_levels))

common_theme <- theme_manuscript(12) +
  theme(
    plot.tag = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 13, colour = "black"),
    axis.text = element_text(size = 11, colour = "black"),
    strip.text = element_text(size = 12, face = "bold", colour = "black"),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 11),
    legend.key.size = grid::unit(0.7, "cm"),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4),
    panel.grid.minor = element_blank(),
    plot.margin = margin(t = 7, r = 9, b = 7, l = 7)
  )

panel_a <- grouped_r2 |>
  mutate(
    framework_short = factor(
      framework,
      levels = c("Lotka-Volterra", "Consumer-resource"),
      labels = framework_levels
    )
  ) |>
  ggplot(aes(
    x = treatment_plot,
    y = r2,
    colour = framework_short,
    fill = framework_short
  )) +
  stat_halfeye(
    position = position_dodge(width = 0.65),
    adjust = 0.8,
    width = 0.85,
    slab_alpha = 0.40,
    point_interval = median_qi,
    .width = c(0.5, 0.8),
    point_size = 3
  ) +
  geom_point(
    position = position_jitterdodge(dodge.width = 0.65, jitter.width = 0.03),
    alpha = 0.20,
    size = 2.2,
    show.legend = FALSE
  ) +
  facet_wrap(~ predictor, nrow = 1) +
  scale_colour_manual(values = framework_colours, name = "Framework", drop = FALSE) +
  scale_fill_manual(values = framework_colours, drop = FALSE, guide = "none") +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    x = "Interaction strength / resource sharing",
    y = expression(paste("Grouped explanatory power (", R^2, ")"))
  ) +
  common_theme +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 11, angle = 0, hjust = 0.5),
    strip.text = element_text(size = 13, face = "bold")
  )

random_subset_panel <- random_subset_data |>
  ggplot(aes(x = r2, fill = experiment_label)) +
  geom_histogram(
    aes(y = after_stat(density)),
    bins = 32,
    colour = "white",
    linewidth = 0.25,
    alpha = 0.75
  ) +
  geom_density(linewidth = 0.9, colour = "grey20") +
  facet_wrap(~ experiment_label, ncol = 2) +
  scale_fill_manual(values = experiment_colours, drop = FALSE, guide = "none") +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = expression(paste("Random-subset ", R^2, " for actual CV(CP)")),
    y = "Density"
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 10.5),
    strip.text = element_text(size = 11.5, face = "bold")
  )

importance_plot <- importance_data |>
  ggplot(aes(
    y = variable_y,
    x = scaled_importance,
    colour = experiment_label,
    shape = experiment_label
  )) +
  geom_segment(
    aes(x = 0, xend = scaled_importance, yend = variable_y),
    linewidth = 0.65
  ) +
  geom_point(size = 2.8, stroke = 0.6) +
  scale_colour_manual(values = experiment_colours, drop = FALSE, name = "Run") +
  scale_shape_manual(values = c(16, 17, 15, 18), drop = FALSE, name = "Run") +
  scale_y_continuous(
    breaks = seq_along(levels(importance_data$variable_label)),
    labels = levels(importance_data$variable_label)
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, by = 0.25)) +
  labs(
    x = "Scaled permutation importance",
    y = "Predictor variables"
  ) +
  common_theme +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 11),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4)
  )

rf_accuracy_panel <- rf_prediction_data |>
  ggplot(aes(x = predicted, y = observed, colour = experiment_label)) +
  geom_point(alpha = 0.25, size = 1) +
  geom_abline(
    slope = 1,
    intercept = 0,
    linetype = "dashed",
    colour = "grey50",
    linewidth = 0.65
  ) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +
  geom_text(
    data = rf_prediction_data |> distinct(experiment_label, oob_label),
    aes(x = -Inf, y = Inf, label = oob_label),
    inherit.aes = FALSE,
    hjust = -0.08,
    vjust = 1.3,
    size = 3.8,
    colour = "black"
  ) +
  facet_wrap(~ experiment_label, ncol = 2, scales = "free") +
  scale_colour_manual(values = experiment_colours, drop = FALSE, guide = "none") +
  labs(
    x = "Out-of-bag predicted instability",
    y = "Observed instability"
  ) +
  common_theme +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10.5),
    axis.text.y = element_text(size = 10.5),
    strip.text = element_text(size = 11.5, face = "bold")
  )

panel_a /
  (random_subset_panel | importance_plot | rf_accuracy_panel) +
  plot_layout(heights = c(1, 1.05), widths = c(1, 1, 1), guides = "keep") +
  plot_annotation(
    tag_levels = "a",
    tag_prefix = "(",
    tag_suffix = ")"
  )
```

Figure 2. Performance of CP-based forecasts in four simulation runs. (a) Grouped \(R^2\) for actual CP variability, null CP variability, and trait summaries across interaction-strength or resource-sharing gradients. Lotka-Volterra CPs use raw performance; consumer-resource CPs use viability. (b) Random-subset robustness. (c) Scaled random-forest permutation importance. (d) Observed versus out-of-bag random-forest predictions.

# Figure 3

```{r figure-3-prep, include=FALSE}
# Load raw data sets
# Load raw data sets
ts_df <- read.csv("~/Documents/GitHub/CPCs_predict_CommunityStability/new_empirical_study/data/combined_timeseries_df.csv")

TPC_timeseries <- read.csv("~/Documents/GitHub/CPCs_predict_CommunityStability/new_empirical_study/data/TPC_timeseries_df.csv")

ts_df_raw <- ts_df
TPC_timeseries_raw <- TPC_timeseries



plot_ts_dens<-ggplot(
  data = TPC_timeseries,
  aes(x = day, y = mean.dens.ml, colour = temperature)
) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_grid(species ~ temperature, scales = "free") +
  scale_colour_viridis_d(option = "magma", end = 0.85)

plot_ts_dens_log<-ggplot(
  data = TPC_timeseries,
  aes(x = day, y = log10(mean.dens.ml), colour = temperature)
) +
  geom_point() +
  geom_smooth(method = "loess", se = FALSE) +
  facet_grid(species ~ temperature, scales="free_y") +
  scale_colour_viridis_d(option = "magma", end = 0.85)


### Fit Thermal performance curves using gams


TPC_dens_df <- TPC_timeseries |> 
  filter(species != "S") |> 
  mutate(
    log_dens = log10(mean.dens.ml ),
    max_days = case_when(
      species == "C" ~ 3,
      species == "D" & temperature %in% c("T24", "T27","T33") ~ 2,
      species == "D" & temperature %in% c("T30") ~ 1,
      species == "D" ~ 3,
      species == "L" & temperature %in% c("T18","T21", "T24", "T27","T30","T33") ~ 2,
      species == "L" ~ 3,
      species == "P" & temperature %in% c("T24", "T27") ~ 2,
      species == "P" & temperature %in% c("T15", "T18") ~ 6,
      species == "P" & temperature %in% c("T30", "T33") ~ 2,
      species == "P" ~ 3,
      species == "T" & temperature %in% c("T30", "T33") ~ 1,
      species == "T" & temperature == "T27" ~ 2,
      species == "T" & temperature %in% c("T18", "T21", "T24") ~ 3,
      species == "T" & temperature == "T15" ~ 4
    )
  ) |> 
  filter(day <= max_days) |> 
  group_by(species, temperature) |> 
  summarise(
    igr = coef(lm(log_dens ~ day))[2],
    n_days = n_distinct(day),
    .groups = "drop"
  ) |> 
  mutate(temperature = as.numeric(sub("T", "", temperature)))

plot_dens_TPCs <- ggplot(
  data = TPC_dens_df,
  aes(x = temperature, y = igr, colour = species)
) +
  geom_point() +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 7),
    se = FALSE
  ) + 
  scale_colour_viridis_d(option = "magma", end = 0.85) +
  theme_bw()



new_data <- data.frame(temperature = seq(15, 33, by= 0.02))

nested_gams <- TPC_dens_df %>%
  nest(cols =-species) %>%
  mutate(
    gams = map(cols, ~ gam(igr ~ s(temperature, k = 7),
                           data = .x,
                           method = "REML")),
    predicted = map(gams, ~ predict(.x, newdata = new_data))
  )

tpc_gams_design <- setNames(nested_gams$gams, nested_gams$species)

predicted <- nested_gams %>% unnest(predicted)
rates <- cbind(new_data, predicted)
rates <- rates %>%
  relocate(species, temperature, predicted)

TPC_predicted_dens<-rates %>% select(species, temperature, predicted)
TPCs <- TPC_predicted_dens

plot_dens_TPCs_predicted <- ggplot(
  data = TPC_predicted_dens,
  aes(x = temperature, y = predicted, colour = species)
) +
  geom_line(size = 2) +
  scale_colour_viridis_d(option = "magma", end = 0.85) +
  theme_bw()


### Calculate Null Community performance variability (CV(CP))

species_list <- c("L", "D", "P", "C", "T")
richness_levels <- c(2, 3, 4)

all_combinations <- map_dfr(richness_levels, function(r) {
  combs <- combn(species_list, r, simplify = FALSE)
  tibble(
    richness = r,
    composition_name = map_chr(combs, ~ paste(sort(.x), collapse = "")),
    species = combs
  )
})

list_combinations<- split(all_combinations,all_combinations$composition_name)

list_CVperf<- lapply(list_combinations,function(r){
  
    community <- r[["species"]][[1]]

  selected <- TPC_predicted_dens %>% 
  dplyr::filter(species %in% community) 

  sum_curve<-selected %>% 
    group_by(temperature) %>% 
    summarise(tot_igr=sum(predicted))
  
  pop_CV<-selected %>% 
    group_by(species) %>% 
    summarise(sd=sd(predicted),
              mean=mean(predicted),
              weight=mean/ mean(sum_curve$tot_igr),
              CV=sd/mean) |> 
    ungroup() |> 
    summarise(pop_CVperf_null =sum(CV*weight),
              synchrony=(sd(sum_curve$tot_igr))^2/(sum(sd))^2)
  
  CV_perf <- sd(sum_curve$tot_igr)/mean(sum_curve$tot_igr)
  return(data.frame(CV_perf_null=CV_perf,
                    pop_CVperf_null=pop_CV$pop_CVperf_null,
                    synchrony_null=pop_CV$synchrony,
                    composition=r["composition_name"],
                    richness=r["richness"]))
    
})

df_CVperf_naive_dens<-do.call("rbind", list_CVperf)

plot_CV_perfs_naive <- ggplot(data = df_CVperf_naive_dens, 
                        aes(x = as.factor(richness), y = CV_perf_null)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(color = as.factor(richness)), 
              width = 0.2, 
              alpha = 0.6) +
  geom_text_repel(aes(label = composition_name),
                  size = 3,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(x = "Species Richness", 
       y = "CV of Performance",
       title = "Community CV by Richness",
       color = "Richness")

### Create Temperature Time Series

make_reddened_temperature <- function(seed) {
  n <- 35
  gamma <- 0.8
  target_mean <- 22.5
  target_sd <- 3.5
  temp_step <- 0.1

  set.seed(seed)
  f <- seq(1, n / 2)
  scales <- 1 / (f^(gamma / 2))
  phases <- runif(length(f), 0, 2 * pi)
  coeffs <- scales * exp(1i * phases)
  full_coeffs <- c(0, coeffs, rev(Conj(coeffs[-length(coeffs)])))
  raw_noise <- Re(fft(full_coeffs, inverse = TRUE))[1:n]
  temperature_reddened <- as.vector(scale(raw_noise)) * target_sd + target_mean
  temperature_reddened <- round(temperature_reddened / temp_step) * temp_step

  data.frame(
    time = 1:n,
    temp = temperature_reddened
  ) %>% 
    drop_na()
}

df_red1 <- make_reddened_temperature(246)
df_red2 <- make_reddened_temperature(113)
df_red3 <- make_reddened_temperature(1356)
df_red4 <- make_reddened_temperature(3356)

df_temp <- tibble(
  time = df_red1$time,
  TS1 = df_red1$temp,
  TS2 = df_red2$temp,
  TS3 = df_red3$temp,
  TS4 = df_red4$temp
)

temp_series <- df_temp |> pivot_longer(cols = starts_with("TS"), values_to = "temp" ,names_to="temp_series")

### Calculate Actual Community performance variability (CV(CP)) 

calculate_actual_cv <- function(T_norm) {
  lapply(list_combinations,function(r){
    
      community <- r[["species"]][[1]]

    selected <- TPC_predicted_dens %>% 
    dplyr::filter(species %in% community, temperature %in% T_norm) 

    sum_curve<-selected %>% 
      group_by(temperature) %>% 
      summarise(tot_igr=sum(predicted))
    
    pop_CV<-selected %>% 
      group_by(species) %>% 
      summarise(sd=sd(predicted),
                mean=mean(predicted),
                weight=mean/ mean(sum_curve$tot_igr),
                CV=sd/mean) |> 
      ungroup() |> 
      summarise(pop_CVperf_actual =sum(CV*weight),
                 synchrony=(sd(sum_curve$tot_igr))^2/(sum(sd))^2)
    
    CV_perf <- sd(sum_curve$tot_igr)/mean(sum_curve$tot_igr)
    return(data.frame(CV_perf_actual=CV_perf,
                      pop_CVperf_actual=pop_CV$pop_CVperf_actual,
                      synchrony_actual=pop_CV$synchrony,
                      composition=r["composition_name"],
                      richness=r["richness"]))
  }) |>
    do.call(what = "rbind")
}

T_norm1 <- df_red1$temp
T_norm1<-T_norm1[c(1:5,8:12,15:19,22:26, 29:33)]
df_CVperf_obs_dens1 <- calculate_actual_cv(T_norm1)

plot_CV_perfs1 <- ggplot(data = df_CVperf_obs_dens1, 
                        aes(x = as.factor(richness), y = CV_perf_actual)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  geom_text_repel(aes(label = composition_name),
                  size = 3,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(x = "Species Richness", 
       y = "CV of Performance Observed",
       title = "1: Community CV by Richness",
       color = "Richness")

comp_choice1 <- df_CVperf_obs_dens1 %>% 
  filter(composition_name %in% c("PT", "CD", "LT", "CLT", "LPT","DPT", "CDLT", "DLPT", "CDLP")) %>% 
  mutate(temp_series = 1)

T_norm2 <- df_red2$temp
T_norm2<-T_norm2[c(1:5,8:12,15:19,22:26, 29:33)]
df_CVperf_obs_dens2 <- calculate_actual_cv(T_norm2)

plot_CV_perfs2 <- ggplot(data = df_CVperf_obs_dens2, 
                        aes(x = as.factor(richness), y = CV_perf_actual)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  geom_text_repel(aes(label = composition_name),
                  size = 3,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(x = "Species Richness", 
       y = "CV of Performance Observed",
       title = "2: Community CV by Richness",
       color = "Richness")

comp_choice2 <- df_CVperf_obs_dens2 %>% 
  filter(composition_name %in% c("LT", "CP", "PT", "CLT", "DLP","DPT", "CDLT", "CDLP", "CDPT")) %>% 
  mutate(temp_series = 2)

T_norm3 <- df_red3$temp
T_norm3<-T_norm3[c(1:5,8:12,15:19,22:26, 29:33)]
df_CVperf_obs_dens3 <- calculate_actual_cv(T_norm3)

plot_CV_perfs3 <- ggplot(data = df_CVperf_obs_dens3, 
                        aes(x = as.factor(richness), y = CV_perf_actual)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  geom_text_repel(aes(label = composition_name),
                  size = 3,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(x = "Species Richness", 
       y = "CV of Performance Observed",
       title = "3: Community CV by Richness",
       color = "Richness")

comp_choice3 <- df_CVperf_obs_dens3 %>% 
  filter(composition_name %in% c("LT", "CD", "PT", "CLT", "DLP","DPT", "CDLT", "CDLP", "CDPT")) %>% 
  mutate(temp_series = 3)

T_norm4 <- df_red4$temp
T_norm4<-T_norm4[c(1:5,8:12,15:19,22:26, 29:33)]
df_CVperf_obs_dens4 <- calculate_actual_cv(T_norm4)

plot_CV_perfs4 <- ggplot(data = df_CVperf_obs_dens4, 
                        aes(x = as.factor(richness), y = CV_perf_actual)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point() +
  geom_text_repel(aes(label = composition_name),
                  size = 3,
                  max.overlaps = 15,
                  box.padding = 0.5,
                  point.padding = 0.3,
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(x = "Species Richness", 
       y = "CV of Performance Observed",
       title = "4: Community CV by Richness",
       color = "Richness")

comp_choice4 <- df_CVperf_obs_dens4 %>% 
  filter(composition_name %in% c("LT", "LP", "PT", "CLT", "LPT","DPT", "CDLT", "CDLP", "CDPT")) %>% 
  mutate(temp_series = 4)

all_comps <- bind_rows(comp_choice1, comp_choice2, comp_choice3, comp_choice4)

### Calcualate Trait summaries

species_T_opt <- TPC_predicted_dens %>% 
  filter(species != "S") %>% 
  group_by(species) %>% 
  slice_max(predicted, n = 1, with_ties = FALSE) %>% 
  ungroup() %>% 
  transmute(species, T_opt = temperature)

composition_T_opt_summary <- all_comps %>% 
  distinct(composition_name, richness) %>% 
  left_join(all_combinations, by = c("composition_name", "richness")) %>% 
  unnest(species) %>% 
  left_join(species_T_opt, by = "species") %>% 
  group_by(composition_name, richness) %>% 
  summarise(
    T_opt_mean = abs(mean(T_opt) - 22.5),
    sd_T_opt = sd(T_opt),
    .groups = "drop"
  )

all_comps <- left_join(all_comps, df_CVperf_naive_dens, by = c("composition_name", "richness"))
all_comps <- left_join(all_comps, composition_T_opt_summary, by = c("composition_name", "richness"))

### Calculate TPC overlap (not used in analysis)

trapz <- function(x, y) {
  sum(diff(x) * (head(y, -1) + tail(y, -1)) / 2)
}

tpc_density_shapes <- TPC_predicted_dens |>
  arrange(species, temperature) |>
  group_by(species) |>
  mutate(
    positive_predicted = pmax(predicted, 0),
    density = positive_predicted / trapz(temperature, positive_predicted)
  ) |>
  ungroup()

calculate_pairwise_tpc_overlap <- function(species_pair) {
  pair_curves <- tpc_density_shapes |>
    filter(species %in% species_pair) |>
    select(species, temperature, density) |>
    pivot_wider(names_from = species, values_from = density) |>
    arrange(temperature)
  curve_names <- species_pair

  trapz(
    pair_curves$temperature,
    pmin(pair_curves[[curve_names[1]]], pair_curves[[curve_names[2]]])
  )
}

calculate_shared_tpc_overlap <- function(community) {
  community_curves <- tpc_density_shapes |>
    filter(species %in% community) |>
    select(species, temperature, density) |>
    pivot_wider(names_from = species, values_from = density) |>
    arrange(temperature)

  trapz(
    community_curves$temperature,
    do.call(pmin, community_curves[community])
  )
}

community_tpc_pairwise_overlap <- all_combinations |>
  mutate(species_pairs = map(species, ~ combn(.x, 2, simplify = FALSE))) |>
  select(composition_name, richness, species_pairs) |>
  unnest_longer(species_pairs) |>
  mutate(
    species_pair = map_chr(species_pairs, ~ paste(sort(.x), collapse = "_")),
    pairwise_tpc_overlap = map_dbl(species_pairs, calculate_pairwise_tpc_overlap)
  ) |>
  select(composition_name, richness, species_pair, pairwise_tpc_overlap)

community_tpc_overlap <- community_tpc_pairwise_overlap |>
  group_by(composition_name, richness) |>
  summarise(
    n_pairs = n(),
    mean_pairwise_tpc_overlap = mean(pairwise_tpc_overlap),
    min_pairwise_tpc_overlap = min(pairwise_tpc_overlap),
    max_pairwise_tpc_overlap = max(pairwise_tpc_overlap),
    sd_pairwise_tpc_overlap = sd(pairwise_tpc_overlap),
    .groups = "drop"
  ) |>
  left_join(
    all_combinations |>
      mutate(shared_tpc_overlap = map_dbl(species, calculate_shared_tpc_overlap)) |>
      select(composition_name, richness, shared_tpc_overlap),
    by = c("composition_name", "richness")
  )

all_comps <- left_join(all_comps, community_tpc_overlap, by = c("composition_name", "richness"))
CPCs <- all_comps

list_species<- split(species_list,species_list)

list_CVperf<- lapply(list_species,function(s){
  selected <- TPC_predicted_dens %>% 
  dplyr::filter(species == s) 
  
  CV_perf <- sd(selected$predicted)/mean(selected$predicted)
  return(data.frame(CV_perf_null_species=CV_perf,species=s))
})

df_CVperf_naive_species<-do.call("rbind", list_CVperf)

calculate_actual_species_cv <- function(T_norm, temp_series_name) {
  lapply(list_species,function(s){
    selected <- TPC_predicted_dens %>% 
    dplyr::filter(species %in% s, temperature %in% T_norm) 
    
    CV_perf <- sd(selected$predicted)/mean(selected$predicted)
    return(data.frame(CV_perf_actual_species=CV_perf,species=s, temp_series=temp_series_name))
  }) |>
    do.call(what = "rbind")
}

df_CVperf_actual_species_TS1 <- calculate_actual_species_cv(df_red1$temp, "TS1")
df_CVperf_actual_species_TS2 <- calculate_actual_species_cv(df_red2$temp, "TS2")
df_CVperf_actual_species_TS3 <- calculate_actual_species_cv(df_red3$temp, "TS3")
df_CVperf_actual_species_TS4 <- calculate_actual_species_cv(df_red4$temp, "TS4")

df_CVperf_actual_species<-rbind(df_CVperf_actual_species_TS1,df_CVperf_actual_species_TS2,df_CVperf_actual_species_TS3,df_CVperf_actual_species_TS4)

df_CVperf_species<-left_join(df_CVperf_actual_species,df_CVperf_naive_species)

# Remove initial stable period 
ts_df <- ts_df %>%
  filter(day>4) %>% 
  mutate(day=day-4)


# Remove only the specific sampling error days.
outlier_sampling_days <- tribble(
  ~sample_ID, ~day,
  "DLP_TS3_2", 4,
  "PT_TS4_1", 1,
  "LP_TS4_1", 3
)

ts_df <- ts_df %>%
  anti_join(outlier_sampling_days, by = c("sample_ID", "day"))


# Create  variables

ts_df <- ts_df %>% mutate(richness=as.factor(nchar(composition_name)))

CPCs<-CPCs %>% mutate(temp_series=paste0("TS",temp_series),
                      richness = as.factor(richness))

# Detrend each species-level time series with one smooth trend pooled across replicates.
# Within each composition x temperature x species, the GAM day trend is fit
# using all three replicate time series, then removed from each observation.


get_pooled_gam_trend <- function(y, day) {
  if (length(y) >= 4 && n_distinct(day, na.rm = TRUE) >= 4) {
    trend_data <- tibble(y = y, day = day)
    smooth_k <- 5 
    trend_model <- gam(y ~ s(day, k = smooth_k), data = trend_data, method = "REML")
    pooled_trend <- predict(trend_model, newdata = trend_data)
  } else {
    pooled_trend <- rep(mean(y, na.rm = TRUE), length(y))
  }

  pooled_trend
}

ts_df <- ts_df %>%
  group_by(composition_name, predict_spec) %>%
  mutate(
    mean.dens.ml_trend = get_pooled_gam_trend(mean.dens.ml, day),
    biomass_trend = get_pooled_gam_trend(biomass, day),
    mean.dens.ml_detrended = mean.dens.ml - mean.dens.ml_trend + mean(mean.dens.ml_trend, na.rm = TRUE),
    biomass_detrended = biomass - biomass_trend 
  ) %>%
  ungroup() 



  

# Calculate community variability 

CV_0_df <- ts_df %>% 
  group_by(sample_ID, composition_name, temp_series, replicate, richness, day) %>% 
  summarise(com_biom=sum(biomass_detrended),
            Para ="P" %in% strsplit(composition_name, split="")[[1]]) %>% 
  ungroup() %>% 
  group_by(sample_ID, composition_name, temp_series, replicate, richness, Para) %>% 
  summarise( mean_biom=mean(com_biom),
            sd_biom=sd(com_biom))



# population variability / synchrony

pop_CV <- ts_df |> 
  group_by(sample_ID, composition_name, temp_series, replicate, richness, predict_spec) |> 
  summarise(sd_pop_biom = sd(biomass_detrended)) |> 
  ungroup() 
pop_CV<-pop_CV |> left_join(CV_0_df, by = c("sample_ID", "composition_name", "temp_series", "replicate", "richness"))



pop_CV <- pop_CV |> 
  group_by(sample_ID, composition_name, temp_series, replicate, richness) |> 
  summarise(mean_sd_pop = mean(sd_pop_biom),
            synchrony = unique(sd_biom)^2/(sum(sd_pop_biom))^2)


CV_0_df <- CV_0_df |> left_join(pop_CV, by = c("sample_ID", "composition_name", "temp_series", "replicate", "richness"))

# Join with CPC

CPC_CV_df <- CV_0_df %>% 
  full_join(CPCs) %>%
  mutate(richness_num = as.numeric(as.character(richness)))



# Create models

m_CPC_obs<-lm(data=CPC_CV_df, log10(sd_biom)~log10(CV_perf_actual))
sum_CPC_obs<-summary(m_CPC_obs)
r2_obs <- sum_CPC_obs$adj.r.squared



MASS::boxcox(lm(data=CPC_CV_df, (sd_biom)~(sd_T_opt)))

m_CPC_naive<-lm(data=CPC_CV_df, log10(sd_biom)~log10(CV_perf_null))

sum_CPC_naive<-summary(m_CPC_naive)
r2_naive <- sum_CPC_naive$adj.r.squared

m_T_opt_mean<-lm(data=CPC_CV_df, log10(sd_biom)~(T_opt_mean))

sum_T_opt_mean<-summary(m_T_opt_mean)
r2_T_opt_mean <- sum_T_opt_mean$adj.r.squared

m_T_opt_sd<-lm(data=CPC_CV_df, log10(sd_biom)~sd_T_opt)

sum_T_opt_sd<-summary(m_T_opt_sd)
r2_T_opt_sd <- sum_T_opt_sd$adj.r.squared

m_traits_interaction <-lm(data=CPC_CV_df, log10(sd_biom)~sd_T_opt*T_opt_mean)

sum_traits_interaction<-summary(m_traits_interaction)
r2_traits_interaction <- sum_traits_interaction$adj.r.squared

### Functions for extracting model parameters

format_p <- function(p) {
  ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
}

get_model_stats <- function(model, predictor_label, predictor_term = 2) {
  model_summary <- summary(model)
  coefficients <- model_summary$coefficients
  predictor_coefficients <- coefficients[predictor_term, ]

  tibble(
    predictor = predictor_label,
    beta = predictor_coefficients["Estimate"],
    t_value = predictor_coefficients["t value"],
    p_value = predictor_coefficients["Pr(>|t|)"],
    r_squared = model_summary$r.squared
  )
}

### Main results

predictor_stats <- bind_rows(
  get_model_stats(m_CPC_obs, "Actual community performance variability"),
  get_model_stats(m_CPC_naive, "Null community performance variability"),
  get_model_stats(m_T_opt_mean, "|mean Topt - mean temperature|"),
  get_model_stats(m_T_opt_sd, "sd(Topt)"),
  get_model_stats(m_traits_interaction, "Trait summaries")
) %>%
  mutate(
    beta_text = formatC(beta, format = "e", digits = 2),
    t_text = sprintf("%.2f", t_value),
    p_text = format_p(p_value),
    r2_text = sprintf("%.2f", r_squared)
  )

obs_stats <- predictor_stats %>% filter(predictor == "Actual community performance variability")
naive_stats <- predictor_stats %>% filter(predictor == "Null community performance variability")
topt_mean_stats <- predictor_stats %>% filter(predictor == "|mean Topt - mean temperature|")
topt_sd_stats <- predictor_stats %>% filter(predictor == "sd(Topt)")
traits_interaction_stats <- predictor_stats %>% filter(predictor == "Trait summaries")

### Results full model for CV(CP_actual)


m_full<-lm(data=CPC_CV_df, log10(sd_biom)~log10(CV_perf_actual)*richness+temp_series)

sum_full<-summary(m_full)
r2_full_adj<-sum_full$adj.r.squared
r2_full_adj_text <- sprintf("%.2f", r2_full_adj)

full_model_f <- unname(sum_full$fstatistic["value"])
full_model_df_num <- unname(sum_full$fstatistic["numdf"])
full_model_df_den <- unname(sum_full$fstatistic["dendf"])
full_model_p <- pf(full_model_f, full_model_df_num, full_model_df_den, lower.tail = FALSE)
full_model_f_text <- sprintf("%.2f", full_model_f)
full_model_p_text <- format_p(full_model_p)

anova_full <- anova(m_full)
full_perf_f <- anova_full["log10(CV_perf_actual)", "F value"]
full_perf_df_num <- anova_full["log10(CV_perf_actual)", "Df"]
full_perf_p <- anova_full["log10(CV_perf_actual)", "Pr(>F)"]
full_perf_f_text <- sprintf("%.2f", full_perf_f)
full_perf_p_text <- format_p(full_perf_p)

full_temp_f <- anova_full["temp_series", "F value"]
full_temp_df_num <- anova_full["temp_series", "Df"]
full_temp_p <- anova_full["temp_series", "Pr(>F)"]
full_temp_f_text <- sprintf("%.2f", full_temp_f)
full_temp_p_text <- format_p(full_temp_p)

full_interaction_f <- anova_full["log10(CV_perf_actual):richness", "F value"]
full_interaction_df_num <- anova_full["log10(CV_perf_actual):richness", "Df"]
full_interaction_p <- anova_full["log10(CV_perf_actual):richness", "Pr(>F)"]
full_interaction_f_text <- sprintf("%.2f", full_interaction_f)
full_interaction_p_text <- format_p(full_interaction_p)

full_coefficients <- as_tibble(sum_full$coefficients, rownames = "term")

richness_slope_2 <- full_coefficients %>%
  filter(term == "log10(CV_perf_actual)") %>%
  pull(Estimate)
richness_slope_3 <- richness_slope_2 + full_coefficients %>%
  filter(term == "log10(CV_perf_actual):richness3") %>%
  pull(Estimate)
richness_slope_4 <- richness_slope_2 + full_coefficients %>%
  filter(term == "log10(CV_perf_actual):richness4") %>%
  pull(Estimate)

richness_slope_text <- paste0(
  "richness 2: β = ", sprintf("%.2f", richness_slope_2),
  "; richness 3: β = ", sprintf("%.2f", richness_slope_3),
  "; richness 4: β = ", sprintf("%.2f", richness_slope_4)
)

richness_interaction_text <- full_coefficients %>%
  filter(str_detect(term, "^log10\\(CV_perf_actual\\):richness")) %>%
  mutate(
    richness_level = str_remove(term, "^log10\\(CV_perf_actual\\):richness"),
    contrast_text = paste0(
      "richness ", richness_level, " vs. richness 2: β = ",
      formatC(Estimate, format = "e", digits = 2),
      ", *p* = ", format_p(`Pr(>|t|)`)
    )
  ) %>%
  pull(contrast_text) %>%
  paste(collapse = "; ")

temp_series_text <- full_coefficients %>%
  filter(str_detect(term, "^temp_series")) %>%
  mutate(
    temp_level = str_remove(term, "^temp_series"),
    contrast_text = paste0(
      temp_level, " vs. TS1: β = ",
      formatC(Estimate, format = "e", digits = 2),
      ", *p* = ", format_p(`Pr(>|t|)`)
    )
  ) %>%
  pull(contrast_text) %>%
  paste(collapse = "; ")
```




```{r}

#| label: fig-3
#| fig-height: 8.8
#| fig-width: 18
#| fig-align: center
#| message: false
#| warning: false





plot_actual <- ggplot(data=CPC_CV_df, aes(x=log10(CV_perf_actual), y=log10(sd_biom)))+
  geom_point(alpha=0.72, size=2.2, color="black")+
  geom_smooth(aes(group = 1), method="lm", linewidth=0.9, alpha=0.18, color="black", fill = "grey70")+
  theme_bw(base_size = 12)+
  # annotate(
  #   "text",
  #   x = -Inf, y = Inf,
  #   label = paste0("R^2 == ", round(r2_obs, 2)),
  #   parse = TRUE,
  #   hjust = -0.3, vjust = 1.2,
  #   size = 5
  # )  +
  ylab(expression(atop("Community variability", log10(sigma[biom]))))+
  xlab(expression(atop("Actual CP variability", log10(CV(CP[actual])))))


plot_null <- ggplot(data=CPC_CV_df, aes(x=log10(CV_perf_null), y=log10(sd_biom)))+
  geom_point(alpha=0.72, size=2.2, color="black")+
  geom_smooth(aes(group = 1), method="lm", linewidth=0.9, alpha=0.18, color="black", fill = "grey70")+
  theme_bw(base_size = 12)+
  # annotate(
  #   "text",
  #   x = -Inf, y = Inf,
  #   label = paste0("R^2 == ", round(r2_naive, 2)),
  #   parse = TRUE,
  #   hjust = -0.3, vjust = 1.2,
  #   size = 5
  # )  +
  ylab(expression(atop("Community variability", log10(sigma[biom]))))+
  xlab(expression(atop("Null CP variability", log10(CV(CP[null])))))

plot_mean_T_opt <- ggplot(data=CPC_CV_df, aes(x=T_opt_mean, y=log10(sd_biom)))+
  geom_point(alpha=0.72, size=2.2, color="black") +
  geom_smooth(aes(group = 1), method="lm", linewidth=0.9, alpha=0.18, color="black", fill = "grey70")+
  # annotate(
  #   "text",
  #   x = -Inf, y = Inf,
  #   label = paste0("R^2 == ", round(r2_T_opt_mean, 2)),
  #   parse = TRUE,
  #   hjust = -0.3, vjust = 1.2,
  #   size = 5
  # )  +
  theme_bw(base_size = 12)+
  ylab(expression(atop("Community variability", log10(sigma[biom]))))+
  xlab(expression(atop("Mean thermal mismatch", group("|", bar(T)[opt] - bar(T), "|"))))


plot_sd_T_opt <- ggplot(data=CPC_CV_df, aes(x=sd_T_opt, y=log10(sd_biom)))+
  geom_point(alpha=0.72, size=2.2, color="black")+
  geom_smooth(aes(group = 1), method="lm", linewidth=0.9, alpha=0.18, color="black", fill = "grey70")+
  theme_bw(base_size = 12)+
  # annotate(
  #   "text",
  #   x = -Inf, y = Inf,
  #   label = paste0("R^2 == ", round(r2_T_opt_sd, 2)),
  #   parse = TRUE,
  #   hjust = -0.3, vjust = 1.2,
  #   size = 5
  # )  +
  ylab(expression(atop("Community variability", log10(sigma[biom]))))+
  xlab(expression(atop("Thermal optima spread", sd(T[opt]))))

r2_df <- tibble::tibble(
  variable = c(
    "CV_perf_actual",
    "CV_perf_null",
    "T_opt_mean",
    "sd_T_opt",
    "traits_interaction"
  ),
  r_squared = c(
    summary(m_CPC_obs)$adj.r.squared,
    summary(m_CPC_naive)$adj.r.squared,
    summary(m_T_opt_mean)$adj.r.squared,
    summary(m_T_opt_sd)$adj.r.squared,
    summary(m_traits_interaction)$adj.r.squared
  )
) |>
  dplyr::mutate(
    variable = factor(
      variable,
      levels = c("traits_interaction", "sd_T_opt", "T_opt_mean", "CV_perf_null", "CV_perf_actual")
    )
  )

r2_plot <- ggplot(
  r2_df,
  aes(x = r_squared, y = variable)
) +
  geom_segment(
    aes(
      x = 0,
      xend = r_squared,
      yend = variable
    ),
    linewidth = 0.8,
    colour = "grey75"
  ) +
  geom_point(
    size = 3.2,
    colour = "black"
  ) +
  geom_text(
    aes(label = sprintf("%.2f", r_squared)),
    nudge_x = 0.035,
    hjust = 0,
    size = 3.4
  ) +
  scale_y_discrete(
    labels = c(
      "CV_perf_actual" =
        expression(
          atop(
            "Actual CP variability",
            log[10](CV(CP[actual]))
          )
        ),
      
      "CV_perf_null" =
        expression(
          atop(
            "Null CP variability",
            log[10](CV(CP[null]))
          )
        ),
      
      "T_opt_mean" =
        expression(
          atop(
            "Mean thermal mismatch",
            group("|", bar(T)[opt] - bar(T), "|")
          )
        ),
      
      "sd_T_opt" =
        expression(
          atop(
            "Thermal optima spread",
            sd(T[opt])
          )
        ),
      
      "traits_interaction" =
        expression(
          atop(
            "Trait interaction model",
            group("|", bar(T)[opt] - bar(T), "|") %*% sd(T[opt])
          )
        )
    )
  ) +
  scale_x_continuous(
    limits = c(0, 0.85),
    breaks = seq(0, 0.8, by = 0.2),
    expand = expansion(mult = c(0, 0.03))
  ) +
  labs(
    x = expression("Adjusted " * R^2),
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(
      size = 10.5,
      colour = "black",
      lineheight = 0.9
    ),
    axis.text.x = element_text(colour = "black"),
    axis.title.x = element_text(
      margin = margin(t = 8)
    )
  )



combined_design <- c(
  area(t = 1, l = 1, b = 1, r = 1),
  area(t = 1, l = 2, b = 1, r = 2),
  area(t = 2, l = 1, b = 2, r = 1),
  area(t = 2, l = 2, b = 2, r = 2),
  area(t = 1, l = 3, b = 2, r = 3)
)

combined_plot <- wrap_plots(
  plot_actual,
  plot_null,
  plot_mean_T_opt,
  plot_sd_T_opt,
  r2_plot,
  design = combined_design,
  widths = c(1, 1, 0.78),
  guides = "collect"
) +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    plot.tag = element_text(size = 16, face = "bold")
  )

combined_plot
```

Figure 3. Empirical predictors of detrended community biomass variability. (a-d) Relationships between community biomass variability and actual $CP$ variability, null $CP$ variability, mean thermal mismatch, and thermal-optimum spread. (e) Adjusted \(R^2\) values for the corresponding single-predictor and trait-summary models.

# Figure 4

```{r figure-4-synchrony-prep, include=FALSE}
synchrony_data <- analysis_data |>
  transmute(
    experiment,
    experiment_label,
    framework,
    framework_label,
    treatment_plot,
    realised_synchrony = as.numeric(synchrony_abundance),
    `Actual TPC synchrony` = as.numeric(synchrony_performance_info),
    `Null TPC synchrony` = as.numeric(synchrony_performance_naive)
  ) |>
  pivot_longer(
    cols = c(`Actual TPC synchrony`, `Null TPC synchrony`),
    names_to = "synchrony_predictor",
    values_to = "predicted_synchrony"
  ) |>
  filter(is.finite(realised_synchrony), is.finite(predicted_synchrony)) |>
  mutate(
    synchrony_predictor = factor(
      synchrony_predictor,
      levels = c("Actual TPC synchrony", "Null TPC synchrony")
    )
  )

safe_sync_r2 <- function(data) {
  if (
    nrow(data) < 4 ||
      sd(data$realised_synchrony, na.rm = TRUE) == 0 ||
      sd(data$predicted_synchrony, na.rm = TRUE) == 0
  ) {
    return(NA_real_)
  }
  summary(lm(realised_synchrony ~ predicted_synchrony, data = data))$r.squared
}

synchrony_treatment_summary <- synchrony_data |>
  group_by(framework, framework_label, treatment_plot, synchrony_predictor) |>
  group_modify(~ tibble(
    n = nrow(.x),
    r2 = safe_sync_r2(.x),
    correlation = cor(.x$realised_synchrony, .x$predicted_synchrony)
  )) |>
  ungroup()

make_sim_synchrony_panel <- function(framework_name, x_axis_label) {
  synchrony_treatment_summary |>
    filter(framework == framework_name) |>
    ggplot(aes(
      x = treatment_plot,
      y = r2,
      colour = synchrony_predictor,
      group = synchrony_predictor
    )) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    scale_colour_manual(values = c("#3B5C7A", "#B56B45"), name = NULL) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(
      title = unique(synchrony_treatment_summary$framework_label[
        synchrony_treatment_summary$framework == framework_name
      ]),
      x = x_axis_label,
      y = expression(paste("Synchrony relationship ", R^2))
    ) +
    theme_manuscript(12) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      legend.position = "bottom"
    )
}

plot_sync_lv <- make_sim_synchrony_panel("Lotka-Volterra", "Interaction strength")
plot_sync_cr <- make_sim_synchrony_panel("Consumer-resource", "Shared-resource use")

plot_sync_empirical <- ggplot(CPC_CV_df, aes(x = synchrony_actual, y = synchrony)) +
  geom_point(alpha = 0.55, size = 2, colour = "black") +
  geom_smooth(method = "lm", se = TRUE, colour = "black", linewidth = 0.9) +
  annotate(
    "text",
    x = -Inf,
    y = Inf,
    label = paste0("R² = ", syn_r2_text),
    hjust = -0.08,
    vjust = 1.35,
    size = 4.2
  ) +
  xlab(expression(atop("Performance synchrony", phi[perf]))) +
  ylab(expression(atop("Biomass synchrony", phi[biom]))) +
  labs(title = "Empirical experiment") +
  theme_manuscript(12) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

```{r}
#| label: fig-4
#| fig-height: 4.6
#| fig-width: 12
#| fig-align: center
#| message: false
#| warning: false

(plot_sync_lv | plot_sync_cr | plot_sync_empirical) +
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "a", tag_prefix = "(", tag_suffix = ")") &
  theme(
    legend.position = "bottom",
    plot.tag = element_text(size = 16, face = "bold")
  )
```

Figure 4. Synchrony links between thermal performance and community dynamics. (a-b) Explanatory power of performance synchrony for realised biomass synchrony in the Lotka--Volterra and consumer--resource simulations, calculated separately across interaction-strength or shared-resource-use levels. (c) Relationship between performance synchrony and biomass synchrony in the empirical ciliate communities.
